---
title: "Standardization of VTE Prophylaxis for Postpartum Patients"
geometry: margin=1in
author:
- name: Daohai Yu, PhD
  email: dyu@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
- name: J Kyle Armstrong, PhD
  orcid: 0000-0002-4269-0121
  email: johnathon.armstrong@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
date: today
date-format: medium
fontsize: 10pt
bibliography:
  - grateful-refs.bib
format:
  html:
    smooth-scroll: true
    theme:
      light: flatly
      dark: darkly
    toc: true
    lot: true
    lof: true
    number-depth: 6
    number-sections: true
    section-divs: true
    toc-location: left
    toc-expand: true
  pdf:
    toc: true
    lot: true
    lof: true
    number-depth: 6
    standalone: true
    number-sections: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
        \usepackage{pdflscape}
---

```{r setup, echo= FALSE, message=FALSE}
knitr::opts_chunk$set(include = TRUE) 
knitr::opts_chunk$set(echo = FALSE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 

library('purrr')
library('dplyr')
library('ggplot2')
library('data.table')

library('gtsummary')
library('purrr')
library('forcats')
library('kableExtra')
library('arsenal')
set.seed(12345)
```

# Raw Data

```{r}
raw_data <- readxl::read_excel(file.path(here::here(),"data","deidentified data.xlsx")) 
```

## Data Introduction

```{r}
#| label: tbl-raw-data-into
#| tbl-cap: "Raw Data Introduction"

DataExplorer::introduce(raw_data) %>%
  tidyr::pivot_longer(everything()) %>%
  rename(data_attribute = name) %>%
  knitr::kable()
```

```{r}
#| label: fig-raw-data-into-plot
#| fig-cap: "Raw Data Introduction Plot"

DataExplorer::plot_intro(raw_data)
```

```{r}
raw_data <- raw_data %>%
  select(-`...8`,-`...32`)
```

\pagebreak

## Missing Data

In @fig-missing-data we have the missing data we have been able to detect thus far. @fig-missing-data-2 is a refined display containing only columns for which there are some missing values.

```{r}
#| label: fig-missing-data
#| fig-cap: Missing Data

library(datasci)

features_percent_miss(raw_data, na_list = c("NULL")) |>
  plot(add.label = FALSE)
```

```{r}
#| label: fig-missing-data-2
#| fig-cap: Missing Data

features_percent_miss(raw_data, 
                      na_list = c("NULL"), 
                      percent_miss = .0001) |>
  plot()
```

\pagebreak

## Initial Summary Statistics

### Initial Cleaning

For an initial pass we have performed the following:

1.  Converted the gestational age (`GA`) from weeks and days to number of days as numeric. Removed `GA` from analysis.
2.  Standardized the `matches` column to be a factor with levels "yes" for 1 and "no" for 0.
3.  Split the `delivery mode` into `delivery_type` which is either "Vaginal" or "C-Section" and `delivery_type_detail` for all other comments regarding the delivery and removed `delivery mode` from analysis. See @tbl-delivery-type-mode-comp for more details.
4.  Replaced missing with "no" in the columns `major risk factors` and `minor risk factors (2 or more)`.

```{r}
cleaner_data <- raw_data %>%
  mutate(
    GA_days = ifelse(
      GA != "NULL",
      as.numeric(stringr::str_extract(GA, "\\d+(?=w)")) * 7 + as.numeric(stringr::str_extract(GA, "\\d+(?=d)")),
      NA
    )
  ) %>%
  relocate(GA_days, .after = GA) %>%
  select(-GA) %>%
  mutate(matches = case_when(
    matches == 1 ~ 'yes', 
    matches == 0 ~ 'no',
    TRUE ~ 'missing')) %>%
  mutate(matches = factor(matches)) %>%
  mutate(delivery_type =
           case_when(
              `delivery mode` %like% 'C-Section' ~ 'C-Section',
              `delivery mode` %like% 'Vaginal' ~ 'Vaginal',
           )) %>%
  mutate(delivery_type_detail = stringr::str_remove_all(`delivery mode`, delivery_type)) %>%
  mutate(delivery_type_detail = stringr::str_remove_all(delivery_type_detail, "\\(")) %>%
  mutate(delivery_type_detail = stringr::str_remove_all(delivery_type_detail, "\\)")) %>%
  relocate(delivery_type, .after = `delivery mode`) %>%
  relocate(delivery_type_detail, .after = delivery_type) %>%
  mutate(
    across(
      c('major risk factors','minor risk factors (2 or more)'),
      \(x){if_else(is.na(x), "no", x)}
    ))  
```

```{r}
#| label: tbl-delivery-type-mode-comp
#| tbl-cap: "Data cleaning for delivery_type and 'delivery mode' comparison"

cleaner_data %>%
  select(`delivery mode`, delivery_type, delivery_type_detail) %>%
  group_by(`delivery mode`, delivery_type, delivery_type_detail) %>%
  summarise(n = n(), .groups = 'keep') %>%
  arrange(desc(n)) %>%
  knitr::kable()
```

```{r}
cleaner_data <- cleaner_data %>%
  select(-`delivery mode`)
```

@tbl-raw-summary-stats-cont contains the mean, standard deviation, median, inner quartile and range for continuous variables by delivery type.

```{r}
#| label: tbl-raw-summary-stats-cont
#| tbl-cap: "Raw Summary Statistics for continuous variables by Delivery Type"

# cleaner_data %>%
#   select(-ID) %>%
#   select(delivery_type, where(is.numeric)) %>%
#   tbl_summary(by = delivery_type) %>%
#   add_overall(last = TRUE) %>%
#   as_kable_extra(longtable=T, booktabs =T) %>%
#   kable_styling(latex_options=c("repeat_header"))

cleaner_data %>%
  select(-ID) %>%
  select(delivery_type, where(is.numeric)) %>%
  tbl_summary(
    by = delivery_type,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c(
   #   "{N_miss}",
      "{mean} ({sd})",
      "{median} ({p25}, {p75})",
      "{median} ({min} - {max})"
    ),
    digits = everything() ~ 1,
    missing = "ifany"
  ) %>%
  add_overall(last = TRUE) %>%
  #add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2))  %>%
  as_kable_extra(longtable=T, booktabs =T) %>%
  kable_styling(latex_options=c("repeat_header"))
```

@tbl-raw-summary-stats-cat contains the frequency and percentage for categorical variables by delivery type.

```{r}
#| label: tbl-raw-summary-stats-cat
#| tbl-cap: "Raw Summary Statistics for categorical variables by Delivery Type"
#| results: 'asis'

# cleaner_data %>%
#   select(-ID) %>%
#   select(-where(is.numeric)) %>%
#   tbl_summary(by = delivery_type) %>%
#   add_overall(last = TRUE) %>%
#   as_kable_extra(longtable=T, booktabs =T) %>%
#   kable_styling(latex_options=c("repeat_header"))

tableby(delivery_type ~ .,
        data = cleaner_data %>%
          select(-ID) %>%
          select(-where(is.numeric)),
        test=FALSE,
        cat.test = "notest"
        ) %>%
  summary() # pfootnote=TRUE
```

```{r}
report_contingency <- function(my_matrix){
  dimnames_my_matrix <- dimnames(my_matrix)
  
  sum_my_matrix <- sum(my_matrix)
  
  percent_matrix <- round((my_matrix/sum_my_matrix)*100,2)
  percent_matrix
  
  report_matrix <- paste0(my_matrix, " (", percent_matrix,"%)")
  
  final_report_matrix <- matrix(report_matrix, 
                                nrow = nrow(my_matrix),  
                                ncol = ncol(my_matrix),
                                dimnames = dimnames_my_matrix)
  return(final_report_matrix)
}
```

\pagebreak

# Data Cleaning

```{r}
single_level_vars <- cleaner_data %>%
  select(where(is.character) | where(is.factor) | where(is.logical) ) %>%
  colnames() %>%
  purrr::map(\(x){
    cleaner_data %>%
      summarise(n_distinct = n_distinct(!!sym(x))) %>%
      mutate(column = x)
    }) %>%
  list_rbind() %>%
  filter(n_distinct == 1) %>%
  pull(column)
```

Using @tbl-raw-summary-stats-cat we have performed the additional data cleaning steps:

1.  Collapsed `race` into 3 categories: "African American/Black", "Hispanic/Other", which will match the original values, and "Other" which will encapsulate all other values.

    -   Other" encapsulates the following values:
        1.  "Asian/Pacific Islander" (1 total)
        2.  "Caucasian/White" (7 total)
        3.  "Chinese" (1 total)
        4.  "Hispanic" (1 total)
        5.  "Hispanic/ Unknown" (6 total)
        6.  "Other Race" (4 total)
        7.  Unknown (6 total)
    -   Alternately, We could collapse all of the Hispanics into a group for a total count of 51, then our "Other/Unknown" would have 19 members total.

2.  Collapsed `ethnicity` into 3 categories: "Hispanic", "Non-Hispanic", and "Other/Unknown".

3.  Standardized `city` names to be "PHILADELPHIA" instead of "PHILA", although we will remove `city` from the remaining of the analysis as there are only 2 patients outside of Philadelphia.

4.  Removed `state` from the analysis as all patients are from Pennsylvania.

5.  Removed `delivery_type_detail` as there are too many levels to be useful at this time.

6.  Removed the `AC` column and replaced it with columns: `Treatment`, `Treatment_Frequency`, `Treatment_Compliance`, and `Treatment_Special_Conditions`.

    a.  Standardized the `Treatment` column to be "Lovenox 40" if the patient was on Lovenox, "SCDs" if the patient was on SCDs, and "Other" if the patient was on something else.
    b.  Created a `Treatment_Frequency` column to indicate how often the treatment was administered.
    c.  Created a `Treatment_Compliance` column to indicate if the patient was compliant with the treatment.
    d.  Created a `Treatment_Special_Conditions` column to indicate if the patient had any special conditions regarding the treatment (e.g. COVID).

7.  The variables `r paste(single_level_vars[1:(length(single_level_vars)-1)], collapse = ", ")` and `r single_level_vars[length(single_level_vars)]` all have only one unique value (0 variance) and have been removed from the data set.

8.  `antepartum hospitalization` was removed from the analysis as there was only 1 patient who had `antepartum hospitalization`.

9.  `congenital heart disease` was removed from the analysis as there was only 2 patients which had a response other than "no".

10. `active cancer` was removed from the analysis as there was only 1 patient who had history of `active cancer`.

11. `diabetes w/ nephropathy` was removed from the analysis as there were only patients who had "suspected pregestational DM" or "T2DM on inslun", we would need to better understand if there is a more meaningful way to group the various levels captured in this variable.

12. `chorio` was collapsed into 2 categories: "Yes" and "No".

13. `PPH + transfusion/etc` was removed from the analysis as there was only one patient who had a value of "yes".

14. `smoker` has been removed from the analysis as there were no patients with a value of "yes", only 2 patients who were previous smokers.

15. `PEC` was collapsed to 2 categories: "Yes" and "No".

16. `IUGR` was removed from the analysis as there were only 4 patients who had a value of "yes".

17. `drug use` was removed from the analysis as there are only 6 patients who had records of drug use.

18. `cHTN` was collapsed into two categories: "Yes" and "No".

19. removed `matches risk assessment?` as these seem to be related to the `matches` column.

20. `received AC?` was recategorized into `received_ac` with two levels, "yes" and "no".

21. removed `Treatment`, `Treatment_Frequency`, `Treatment_Compliance`, and `Treatment_Special_Conditions` for the time being.

22. `meets_criteria` was added as a new variable to indicate if the patient met the criteria for pharmacologic venous thromboembolism (VTE) prophylaxis. The patient meets the criteria for met criteria for pharmacologic VTE prophylaxis and and `meets_criteria` is defined as `yes` if any of the following are true:

    i.  The patient delivered via C-Section
    ii. The patient has `major risk factors`
    iii. The patient has `minor risk factors (2 or more)`

    -   Otherwise the patient does not meet the criteria and `meets_criteria` is defined as `no`.

```{r}
a_data <- cleaner_data %>%
  mutate(delivery_type = factor(delivery_type)) %>%
  mutate(delivery_type_detail = factor(delivery_type_detail)) %>%
  mutate(city = if_else(city == "PHILA", "PHILADELPHIA", city)) %>%
  select(-all_of(single_level_vars)) %>%
  mutate(across(where(is.character), \(x){factor(x)})) %>%
  mutate(race2 = race) %>%
  mutate(race = fct_lump_n(race2, n=2)) %>%
  select(-race2) %>%
  mutate(ethnicity = fct_lump_n(ethnicity, n=2, other_level = "Other/Unknown")) 
```

```{r}
a_data <- a_data %>%
  mutate(Treatment = 
           case_when(
             grepl('lovenox', AC, ignore.case = TRUE) ~ 'Lovenox 40',
             grepl('SCDs', AC, ignore.case = TRUE) ~ 'SCDs',
             TRUE ~ 'Other')) %>%
  mutate(Treatment_Frequency =
           case_when(
             grepl('BID', AC, ignore.case = TRUE) ~ 'BID',
             grepl('daily', AC, ignore.case = TRUE) ~ 'daily',
             TRUE ~ "NA"
           )) %>%
  mutate(Treatment_Compliance = 
           case_when(
             grepl('refused some doses', AC, ignore.case = TRUE) ~ 'refused_some',
             grepl('refused all doses', AC, ignore.case = TRUE) ~ 'refused_all',
             grepl('lovenox', AC, ignore.case = TRUE) ~ 'compliant',
             TRUE ~ "NA")) %>%
  mutate(Treatment_Conditions =
           case_when(
             grepl('COVID', AC, ignore.case = TRUE) ~ 'COVID',
             TRUE ~ 'none'))
```

```{r}
a_data <- a_data %>% 
  select(-AC) %>%
  mutate(recieved_ac = if_else(`received AC?` == "yes", "yes", "no")) %>%
  relocate(recieved_ac, .after = `received AC?`) %>%
  select(-city,
         -delivery_type_detail,
         -`antepartum hospitalization`,
         -`congenital heart disease`,
         -`active cancer`,
         -`diabetes w/ nephropathy`,
         -`PPH + transfusion/etc`,
         -smoker,
         -IUGR,
         -`drug use`,
         -`received AC?`,
         -`matches risk assessment?`,
         -Treatment, 
         -Treatment_Frequency,
         -Treatment_Compliance,
         -Treatment_Conditions) %>%
  mutate(cHTN = if_else(cHTN =="no","no","yes")) %>%
  mutate(PEC = if_else(cHTN =="no","no","yes")) %>%
  mutate(chorio = if_else(chorio == "no", "no", "yes")) %>%
  mutate(meets_criteria = 
           if_else(
             `major risk factors` == 'yes' | 
             `minor risk factors (2 or more)` == 'yes' | 
              delivery_type == 'C-Section', 'yes', 'no')
         ) %>%
  mutate(across(where(is.character), \(x){factor(x)})) %>%
  mutate(across(where(is.logical), \(x){factor(x)})) 
```

## Summary Statistics by delivery type

```{r}
two_level_vars <- a_data %>%
  select(where(is.character) | where(is.factor) | where(is.logical) ) %>%
  colnames() %>%
  purrr::map(\(x){
    a_data %>%
      summarise(n_distinct = n_distinct(!!sym(x))) %>%
      mutate(column = x)
    }) %>%
  list_rbind() %>%
  filter(n_distinct == 2) %>%
  pull(column)

two_level_vars <- setdiff(two_level_vars, 'delivery_type')
```

```{r}
a_data <- a_data %>%
  #select(all_of(two_level_vars)) %>%
  mutate(
    across(
      setdiff(two_level_vars, 'Treatment'),
      \(x){
        fct_relevel(x, "no")
      }
    )) 
```

Below in @tbl-table4 we have the summary statistics stratified by the delivery type

-   for the categorical covariates with more than 2 levels the $\chi^2$ test was used to test for differences between the groups and Fisher's exact test was used when the expected cell count was less than 5. For the categorical covariates with exactly 2 levels, the mid-p version of Central Fisher's exact test was used to test for differences between the groups.

-   for the continuous data where the Wilcoxon rank sum test was used to test for differences between the C-Section and Vaginal delivery groups.

```{r}
multi_class_levels_vars <- a_data %>%
  select(-ID) %>%
  select(delivery_type, race, ethnicity) %>%
  tbl_summary(
    by = delivery_type) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2))  %>%
  separate_p_footnotes()  %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% # \n N = {n} ({style_percent(p)}%)
  add_overall(last = TRUE, col_label = "**Overall**") #  \n N = {N}
```

```{r}
continuous_vars <- a_data %>%
  select(-ID) %>%
  select(delivery_type, age, BMI, GA_days) %>%
  tbl_summary(
    by = delivery_type,
    type = all_continuous() ~ "continuous2",
    statistic = all_continuous() ~ c(
   #   "{N_miss}",
      "{mean} ({sd})",
      "{median} ({p25}, {p75})",
      "{median} ({min} - {max})"
    ),
    digits = everything() ~ 1,
    missing = "ifany"
  ) %>%
  #add_ci() %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2))  %>%
  add_overall(last = TRUE, col_label = "**Overall**") %>%  #  \n N = {N}    
  separate_p_footnotes() 
```

```{r}
# use this for p-values
custom_exact2x2 <- function(data, variable, by, ...) {
  table(data[[variable]] , data[[by]]) %>%
    exact2x2::exact2x2(midp=T) %>%
    broom::tidy()
}
```

```{r}
# use this for confidence intervals
custom_exact2x2_CI <- function(data, variable, by, ...){
  table(data[[variable]] , data[[by]]) %>%
    exact2x2::exact2x2(midp=T) %>%
    broom::tidy() %>%
  mutate(
    estimate  = style_number(estimate , digits = 1),
    conf.low = style_number(conf.low, digits = 1),
    conf.high = style_number(conf.high, digits = 1)
  ) %>%
  mutate(
    OR_CI = paste0(estimate, " (", conf.low, ", ", conf.high, ")")
  ) %>%
  pull(OR_CI)
}
```

```{r}
binary_vars <- a_data %>%
  select(-ID) %>%
  select(where(is.factor)) %>%
  select(delivery_type, all_of(two_level_vars)) %>%
  tbl_summary(by = delivery_type,
              type = list(all_categorical()~ "categorical")) %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_p(test = all_categorical() ~ "custom_exact2x2") %>% 
  add_overall(last = TRUE, col_label = "**Overall**") %>%  #\n N = {N}      
  separate_p_footnotes() 
```

```{r}
library('gt')
```

```{r}
#| label: tbl-table4
#| tbl-cap: "Summary Statistics stratified by Delivery Type"

tbl_stack(list(multi_class_levels_vars, binary_vars, continuous_vars)) %>% 
  as_gt() %>%
  gt::as_latex()
```

\pagebreak

# Outcome Analysis

```{r}
#| label: tbl-outcome-overall
#| tbl-cap: "Overall Outcome Analysis"

T_final <- a_data %>%
  select(delivery_type, recieved_ac, meets_criteria, matches) %>%
  tbl_summary(by = delivery_type,
              statistic = all_categorical() ~ "{n}/{N} ({p}%)",
              type = list(all_categorical()~ "categorical")
              ) %>%
  add_ci(
    method = everything() ~ "exact",
    statistic = list(all_categorical() ~ "({conf.low}%, {conf.high}%)")) %>%
  add_p(test = all_categorical() ~ "custom_exact2x2") %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_overall(last = TRUE, col_label = "**Overall**") %>% # \n N = {N}
  separate_p_footnotes() %>%
  as_gt() 

T_final %>%
  gt::as_latex()
```

For instance, in @tbl-outcome-overall above, we can see that `r T_final[['_data']] %>% filter(variable == 'meets_criteria') %>% filter(label == 'yes') %>% pull(stat_2)` of the patients who had a vaginal delivery met the criteria for pharmacologic venous thromboembolism (VTE) prophylaxis, with the 95% confidence interval for this proportion being `r T_final[['_data']] %>% filter(variable == 'meets_criteria') %>% filter(label == 'yes') %>% pull(ci_stat_2)`. Whereas, `r T_final[['_data']] %>% filter(variable == 'meets_criteria') %>% filter(label == 'yes') %>% pull(stat_1)` of the patients who had a Cesarean delivery met the criteria for pharmacologic VTE prophylaxis, with the 95% confidence interval for this proportion being `r T_final[['_data']] %>% filter(variable == 'meets_criteria') %>% filter(label == 'yes') %>% pull(ci_stat_1)`. The p-value from the mid-p version of Central Fisher’s Exact Test for the difference between the two groups is `r T_final[['_data']] %>% filter(variable == 'meets_criteria') %>% filter(label != 'no') %>% filter(label != 'yes') %>% pull(p.value)`, which indicates that there is a statistically significant difference between the two groups.

### Meets Criteria for Pharmacologic VTE Prophylaxis 

```{r}
#| label: tbl-outcome-meets-criteria-vaginal
#| tbl-cap: "Outcome Analysis for Meets Criteria for Pharmacologic VTE Prophylaxis in Vaginal Deliveries"

a_data %>%
  filter(delivery_type == "Vaginal") %>%
  select(meets_criteria, matches) %>%
  tbl_summary(by = meets_criteria,
              statistic = all_categorical() ~ "{n}/{N} ({p}%)",
              type = list(all_categorical()~ "categorical")
              ) %>%
  add_ci(
    method = everything() ~ "exact",
    statistic = list(all_categorical() ~ "({conf.low}%, {conf.high}%)")) %>%
  add_p(test = all_categorical() ~ "custom_exact2x2") %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_overall(last = TRUE, col_label = "**Vaginal**") %>% # \n N = {N}
  separate_p_footnotes() %>%
  as_gt() %>%
  gt::as_latex()
```

```{r}
#| label: tbl-outcome-meets-criteria-cesarean
#| tbl-cap: "Outcome Analysis for Meets Criteria for Pharmacologic VTE Prophylaxis in Cesarean Deliveries"

a_data %>%
  filter(delivery_type == "C-Section") %>%
  select(meets_criteria, matches) %>%
  tbl_summary(by = meets_criteria,
              statistic = all_categorical() ~ "{n}/{N} ({p}%)",
              type = list(all_categorical()~ "categorical")
              ) %>%
  add_ci(
    method = everything() ~ "exact",
    statistic = list(all_categorical() ~ "({conf.low}%, {conf.high}%)")) %>%
  add_p(test = all_categorical() ~ "custom_exact2x2") %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_overall(last = TRUE, col_label = "**C-Section**") %>% # \n N = {N}
  separate_p_footnotes() %>%
  as_gt() %>%
  gt::as_latex()
```


```{r}
#| label: tbl-outcome-meets-criteria-overall
#| tbl-cap: "Outcome Analysis for Meets Criteria for Pharmacologic VTE Prophylaxis Overall"

a_data %>%
  select(meets_criteria, matches) %>%
  tbl_summary(by = meets_criteria,
              statistic = all_categorical() ~ "{n}/{N} ({p}%)",
              type = list(all_categorical()~ "categorical")
              ) %>%
  add_ci(
    method = everything() ~ "exact",
    statistic = list(all_categorical() ~ "({conf.low}%, {conf.high}%)")) %>%
  add_p(test = all_categorical() ~ "custom_exact2x2") %>%
  modify_header(all_stat_cols() ~ "**{level}**") %>% #\n N = {n} ({style_percent(p)}%)
  add_overall(last = TRUE, col_label = "**Overall**") %>% # \n N = {N}
  separate_p_footnotes() %>%
  as_gt() %>%
  gt::as_latex()
```


\pagebreak

# Software

```{r}
grateful::cite_packages(output = "paragraph",
                        out.dir = here::here(),
                        omit = NULL,
                        include.RStudio = F,
                        passive.voice = T)
```

\pagebreak

# Refrences
