---
title: "Exploratory Data Analysis, Missingness Auditing, and Normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Exploratory Data Analysis, Missingness Auditing, and Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dev = "png"
)
```

## Overview

In clinical trials, electronic health records (EHR), and epidemiological
registries, data cleanliness is rarely a given. Datasets frequently exhibit
complex missingness patterns, non-standard missing codes (e.g. `"999"`, `"-99"`,
`"Unknown"`, `"Refused"`, `""`), and extreme statistical outliers from data-entry
errors or severe disease phenotypes.

Standard data science tools often assume missingness is cleanly represented as
`NA`, requiring destructive multi-pass preprocessing before exploratory analysis
can even begin. The **`TempleCBE`** package provides domain-tailored biostatistical
tools to audit non-standard missingness non-destructively, rank feature completeness,
standardize numerical features, classify clinical outliers under formal IQR
standards, and transition seamlessly into tuned machine-learning imputation.

```{r setup}
library(TempleCBE)
library(dplyr)
```

---

## Comparison with existing R EDA & missingness tools

The R ecosystem offers several packages for exploratory data analysis and
missingness visualization, notably [naniar](https://naniar.njtierney.com/),
[visdat](https://docs.ropensci.org/visdat/), [mice](https://amices.org/mice/),
and general EDA toolkits like `DataExplorer` and `summarytools`. Understanding
where `TempleCBE` fits alongside these frameworks clarifies its value for
clinical biostatisticians and data managers.

### Existing tools and their limitations in clinical workflows

- **`naniar` & `visdat`**: Pioneered tidy representation and visualization of
  missing data (`vis_miss()`, `gg_miss_var()`). However, they primarily operate
  on existing `NA` values. When raw clinical data contains sentinel values (such
  as `"999"`, `"-99"`, or blank strings), users must first apply mutating
  functions like `replace_with_na_all()`, which can accidentally alter legitimate
  numerical values, modify factor levels, or mutate data types.
- **`mice` / `Amelia`**: Primarily multiple imputation engines. While they
  include basic diagnostics (e.g. `md.pattern()`), their tools are oriented
  toward parameter checking rather than automated feature-level reporting for
  regulatory deliverables.
- **Base R / `dplyr` summaries**: Simple calls like `colSums(is.na(df))` provide
  quick counts, but lack built-in percentage calculations, descending order
  sorting, sentinel code awareness, and publication-ready formatting.

### Where `TempleCBE` provides distinct value

`TempleCBE` was built specifically for the constraints of institutional
biostatistics core facilities:

1. **Zero-mutation non-standard code auditing**: `SumNa()` and
   `features_percent_miss()` natively accept vectors of custom missing codes
   (`na_list = c("999", "-99", "", "Unknown", "NA")`). They compute exact
   missingness counts and percentages *without altering or coercing the underlying
   data frame*.
2. **Two-tier clinical outlier governance**: While generic outlier functions often
   rely on simple z-score cutoffs or arbitrary quantiles, `detect_outliers()`
   implements Tukey's formal biostatistical standard, explicitly distinguishing
   between **MILD** ($1.5 \times \text{IQR}$) and **EXTREME** ($3.0 \times \text{IQR}$)
   outliers and returning structured classification flags that preserve patient keys.
3. **Seamless transition to tuned imputation**: `TempleCBE` links missingness
   auditing directly to high-performance imputation engines (`missforest_mtry()`
   and `missranger_mtry()`) that automatically optimize random forest hyperparameters
   (`mtry`) for the specific missingness rate of the clinical cohort.
4. **Controlled imputation benchmarking**: The `add_missing()` function allows
   statisticians to simulate controlled Missing Completely at Random (MCAR)
   mechanisms on complete datasets to empirically validate downstream imputation
   performance before deploying models.

### Feature comparison matrix

| Dimension | Base / Tidyverse | `naniar` / `visdat` | `TempleCBE` |
|:---|:---|:---|:---|
| **Multi-code missingness** | Manual recoding required | Requires mutating `replace_with_na()` | Built-in via `na_list` argument (zero mutation) |
| **Feature-level ranking** | Multi-line `summarise()` | `miss_var_summary()` | `features_percent_miss()` (sorted tibble + S3 plot) |
| **Missingness heatmap** | Not built-in | `vis_miss()` | `missmap()` (custom clinical palette) |
| **Outlier classification** | Manual IQR calculations | Not covered | Two-tier Tukey standard (MILD: 1.5×, EXTREME: 3.0×) |
| **Feature normalization** | `scale()` (matrix output) | Not covered | `min_max_norm()`, `z_norm()`, `range_norm()` |
| **Imputation handoff** | External | External | Integrated `missforest_mtry()` / `missranger_mtry()` |
| **Controlled amputation** | External (`mice::ampute`) | Limited | `add_missing()` with tracked cell coordinates |

---

## Specific gaps addressed by `TempleCBE`

### 1. The "Dirty Intake" gap in clinical registries
Clinical data extracted from Electronic Data Capture (EDC) platforms like REDCap
or hospital registries frequently blends numeric sentinel values (`999`, `-99`)
and text placeholders (`"Not Recorded"`, `"Declined"`, `""`). Recoding these
in place often converts columns from numeric to character, or loses the ability
to distinguish why data was uncollected. `TempleCBE` audits all codes in their
native types without destructive modifications.

### 2. The "Clinical Outlier Governance" gap
In clinical trials, an extreme lab measurement (e.g. a blood glucose of 650 mg/dL)
might represent a life-threatening clinical event rather than a measurement error.
Blindly trimming or imputing values without clinician review violates regulatory
guidelines (FDA/ICH E9). `detect_outliers()` provides a transparent audit table
flagging both mild and extreme outliers, allowing biostatisticians and medical
monitors to audit values systematically.

### 3. The "Imputation Hyperparameter" gap
Random forest imputation (`missForest` / `missRanger`) is widely considered the gold
standard for mixed clinical data with non-linear relationships. However, analysts
routinely accept default `mtry` settings, which can lead to suboptimal imputation
or excessive computational runtimes on large patient cohorts. `TempleCBE` integrates
automated `mtry` grid search directly into the imputation pipeline.

---

## 1. Auditing Missing Data

### Total missingness with `SumNa()`

`SumNa()` evaluates total missing cells across an entire vector, data frame, or
matrix. When working with raw registry extracts, specify custom sentinel codes
in `na_list`:

```{r sum_na_demo}
# Sample vector with default and non-standard missing codes
vec <- c(12, 14, NA, 999, 18, "", "Unknown")

# Count standard NAs only
SumNa(vec)

# Count NAs and sentinel codes simultaneously
SumNa(vec, na_list = c("999", "", "Unknown"))
```

### Feature-level missingness table with `features_percent_miss()`

In clinical study preparation, biostatisticians need a sorted summary showing
which variables exceed missingness thresholds (e.g. variables with $>20\%$
missingness that cannot be reliably imputed):

```{r features_percent_miss_demo}
# Synthetic clinical cohort
df_clinical <- data.frame(
  patient_id  = 1:6,
  age         = c(45, 52, NA, 61, 38, NA),
  systolic_bp = c(120, 135, NA, 140, 118, 125),
  bmi         = c(22.5, NA, 29.1, NA, 31.0, NA),
  smoker      = c("No", "Yes", "", "No", "NA", "Yes"),
  hba1c       = c(5.6, 999, 6.2, 7.1, 999, 5.8),
  stringsAsFactors = FALSE
)

# Audit missingness including blank strings and sentinel '999' codes
miss_tbl <- features_percent_miss(df_clinical, na_list = c("", "NA", 999))
miss_tbl
```

### Visualizing missingness with `plot_features_percent_miss()` and `missmap()`

`plot_features_percent_miss()` (or its S3 method `plot(miss_tbl)`) generates
a clean horizontal bar chart sorted by missingness frequency:

```{r plot_miss_demo, eval = FALSE}
# Plot missingness across all features
plot_features_percent_miss(df_clinical, na_list = c("", "NA", 999))

# Isolate only the top 3 most missing variables
plot_features_percent_miss(df_clinical, top_n = 3, na_list = c("", "NA", 999))
```

#### Heatmaps with `missmap()`: Row-level and Group-level views

The `missmap()` function visualizes co-occurrence and missingness clustering across
the dataset:

1. **Per-row / Per-column view**: Visualizes each observation individually,
   sorting rows and features by descending missingness count when `row_order = FALSE`:

```{r missmap_row_demo, eval = FALSE}
# Render full patient-level missingness heatmap
missmap(df_clinical, na_list = c("", "NA", 999))
```

2. **Group-aggregated view (`by_column = ...`)**: In multi-center clinical trials,
   auditing missingness aggregated by hospital site, study arm, or visit number
   is critical for monitoring trial conduct:

```{r missmap_grouped_demo, eval = FALSE}
# Synthetic multi-site clinical cohort
df_multisite <- df_clinical |>
  mutate(site = c("Hospital_A", "Hospital_A", "Hospital_B", "Hospital_B", "Hospital_C", "Hospital_C"))

# Aggregate missingness counts per site across clinical features
# Displays a continuous gradient from black to Temple cherry
missmap(df_multisite, by_column = site, na_list = c("", "NA", 999), fill = "count")

# Or display binary presence/absence per site
missmap(df_multisite, by_column = site, na_list = c("", "NA", 999), fill = "binary")
```

---

## 2. Feature Normalization

Machine learning models and penalized regressions require predictors to be
scaled comparably. `TempleCBE` provides three straightforward vector transformations:

- **`min_max_norm()`**: Scales values strictly into the bounded interval $[0, 1]$.
- **`z_norm()`**: Centers values to mean $\mu = 0$ and scales to standard deviation $\sigma = 1$.
- **`range_norm()`**: Scales relative to combined distribution bounds across features.

```{r norm_demo}
vals <- c(10, 20, 30, 40, 50)

# Min-Max normalization [0, 1]
min_max_norm(vals)

# Z-Score standardization (mean = 0, sd = 1)
z_norm(vals)
```

---

## 3. Two-Tier Outlier Detection

`detect_outliers()` audits numerical vectors or data frame columns using Tukey's
interquartile range criteria:
- **Mild Outliers**: Values lying outside $[Q_1 - 1.5 \times \text{IQR}, Q_3 + 1.5 \times \text{IQR}]$.
- **Extreme Outliers**: Values lying outside $[Q_1 - 3.0 \times \text{IQR}, Q_3 + 3.0 \times \text{IQR}]$.

```{r outlier_demo}
clinical_labs <- c(12, 14, 15, 14, 16, 15, 14, 45, 120)

# Detect mild and extreme anomalies
outlier_report <- detect_outliers(clinical_labs)
outlier_report
```

When evaluated on data frames, `detect_outliers()` returns a structured summary
detailing outlier counts, threshold fences, and row indices for clinical auditing.

---

## 4. Complementary Biostatistical Tools in `TempleCBE`

In addition to missingness and scaling, `TempleCBE` provides statistical auditing
utilities that bridge early EDA with formal statistical testing:

### Normality testing and distribution plots

Before applying parametric tests (e.g. Student's t-test or ANOVA), assess distributional
assumptions using `distribution_test()` and `distribution_plot()`:

```{r dist_test_demo}
# Sample biomarker data
biomarker <- c(rnorm(50, mean = 25, sd = 4), 65, 72)

# Conduct Anderson-Darling and Shapiro-Wilk tests
distribution_test(biomarker)
```

```{r dist_plot_demo, eval = FALSE}
# Visual inspection with density overlay and Q-Q plots
distribution_plot(biomarker, feature_name = "Serum Biomarker (ng/mL)")
```

### Automated correlation matrices with `corr_test_all()`

Screen for multi-collinearity across high-dimensional clinical features:

```{r corr_demo, eval = FALSE}
# Compute pairwise correlations with p-values and confidence intervals
corrs <- corr_test_all(df_clinical |> select(age, systolic_bp, bmi))

# Visualize with correlation heatmap
correlation_plot(corrs)
```

### Controlled imputation benchmarking with `add_missing()`

When developing and validating statistical pipelines, `add_missing()` injects
known MCAR missingness into complete datasets:

```{r add_missing_demo}
set.seed(42)
complete_data <- iris[, 1:4]

# Mask 20% of values across selected features
amputed <- add_missing(complete_data, cols = c(Sepal.Length, Petal.Length), pct_na = 0.20)

# Inspect resulting missingness
features_percent_miss(amputed)

# The 'missing_cells' attribute tracks exact amputed coordinates for scoring
head(attr(amputed, "missing_cells"))
```

## 5. Tuned Random Forest Imputation (`missForest` & `missRanger`)

Once missingness patterns have been audited, biostatisticians must decide how to
handle missing values. Complete-case analysis discards valuable clinical data
and introduces bias. Random forest imputation is widely considered the state of
the art for mixed clinical data because it accommodates non-linear relationships
and complex interactions without requiring parametric distribution assumptions.

### The `mtry` dilemma in clinical data

Random forest imputation iterates through each variable with missing values,
fitting a forest predicting that variable from all other variables. The single
most impactful hyperparameter is **`mtry`** — the number of candidate variables
sampled at each split:
- If `mtry` is too low, strong predictive biomarkers may be missed at key splits.
- If `mtry` is too high, trees become correlated and overfit noisy clinical features.

Critically, **different clinical variables perform best at different `mtry` values**.
A continuous biomarker (e.g. serum creatinine) may achieve minimal error at $mtry = 2$,
whereas a multi-level categorical diagnosis (e.g. ICD-10 disease subtype) may
require $mtry = 5$. Standard implementations force the analyst to choose a single
global `mtry` for the entire dataset.

### The `TempleCBE` solution: Independent per-column best `mtry` assembly

`TempleCBE` provides two complementary swept imputation engines:
- **`missforest_sweep_mtry()`**: Based on the gold-standard [missForest::missForest()]
  algorithm.
- **`missranger_sweep_mtry()`**: High-performance alternative powered by [missRanger::missRanger()]
  and the C++ `ranger` library, with support for Predictive Mean Matching (PMM).

Both functions follow a unified, mathematically principled workflow:
1. **Sweep**: Fit the forest across a grid of candidate `mtry` values (e.g. `1:(p - 1)`).
2. **Score**: Track variable-wise out-of-bag (OOB) error for every column at every `mtry`.
3. **Assemble (`assemble_by_best_mtry`)**: Pull each column independently from the
   specific run where that column achieved its minimal OOB error!
4. **Guardrails**: Exclude identifiers (`exclude`), enforce sparsity limits (`max_pct_missing`),
   and refuse all-`NA` columns (`check_no_all_na_columns`).

### High-accuracy imputation with `missforest_sweep_mtry()`

`missforest_sweep_mtry()` reports raw Mean Squared Error (MSE) for numeric columns
and Proportion of Falsely Classified (PFC) for categorical factors:

```{r missforest_sweep_demo, eval = FALSE}
# Synthetic clinical data with missing values
clinical_data <- data.frame(
  patient_id = paste0("PT_", 1:12),
  age        = c(45, 52, NA, 61, 38, 59, 63, NA, 49, 55, 71, 42),
  systolic_bp= c(120, 135, NA, 140, 118, 125, 138, 144, NA, 130, 150, 122),
  bmi        = c(22.5, NA, 29.1, NA, 31.0, 26.4, 28.2, 33.1, 24.8, NA, 27.5, 23.9),
  smoker     = factor(c("No", "Yes", "No", "No", "Yes", "Yes", "No", "Yes", "No", "No", "Yes", "No")),
  stage      = factor(c("I", "II", "II", "III", "I", "II", "IV", "III", "I", "II", "III", "I"))
)

# Impute with mtry sweep (1 to 4), excluding patient_id
res_forest <- missforest_sweep_mtry(
  data            = clinical_data,
  exclude         = "patient_id",
  mtry_values     = 1:3,
  ntree           = 100,
  max_pct_missing = 0.50,
  parallel        = FALSE
)

# 1. Inspect the completed dataset (patient_id reattached in original order)
head(res_forest$imp_data)

# 2. Inspect the winning mtry per column
res_forest$best

# 3. View the full OOB error grid across swept mtry values
res_forest$oob_error
```

### Fast, large-scale imputation with `missranger_sweep_mtry()`

For large clinical cohorts (e.g. $> 10,000$ patients or $> 100$ variables),
`missForest` can be computationally intensive. `missranger_sweep_mtry()` leverages
the C++ `ranger` implementation for high speed:

- **Scaled OOB error**: Reports $1 - R^2$ for numeric columns (where values near 0
  indicate near-perfect recovery and values near 1 indicate performance no better
  than predicting the mean).
- **Predictive Mean Matching (`pmm.k`)**: Setting `pmm.k > 0` (e.g. `pmm.k = 3`)
  ensures that imputed values are drawn from actual observed patient donor values
  rather than synthetic regression averages — critical for bounded clinical scores
  (e.g. integer counts, survey scores).
- **Parallel evaluation**: Uses `furrr` to evaluate candidate `mtry` values across
  CPU cores in parallel without thread oversubscription (`num.threads = 1` inside
  `ranger`).

```{r missranger_sweep_demo, eval = FALSE}
library(furrr)
plan(multisession, workers = 2)

# Fast sweep with predictive mean matching
res_ranger <- missranger_sweep_mtry(
  data        = clinical_data,
  exclude     = "patient_id",
  mtry_values = 1:3,
  num.trees   = 200,
  pmm.k       = 3,
  seed        = 2026,
  parallel    = TRUE
)

# Inspect winning mtry values from ranger
res_ranger$best
```

### Safety and data governance guardrails

Both sweep functions incorporate essential clinical data safeguards:
1. **Identifier and timestamp protection (`exclude`)**: Clinical metadata (patient
   IDs, specimen barcodes, visit dates) are quarantined from the forest training
   matrix and re-attached unaltered to the final output.
2. **Preventing data fabrication (`max_pct_missing`)**: Features that are $>50\%$
   missing (or any custom threshold) are automatically held out from imputation,
   carried through unimputed, and listed in `excluded_high_missing`.
3. **Refusing all-`NA` columns (`check_no_all_na_columns`)**: Standard `missForest`
   silently drops all-`NA` columns, while `missRanger` silently returns them
   still `NA`. `TempleCBE` explicitly halts and warns the user so schema integrity
   is never compromised.

---

## How others can use this technology (Adoption Patterns)

Here are three common patterns for applying `TempleCBE`'s EDA tools to real-world
studies:

### Pattern 1: Clinical trial data intake audit

Create a standardized data quality intake report for newly exported registry data:

```{r pattern_intake, eval = FALSE}
audit_clinical_intake <- function(raw_df) {
  message("Auditing Clinical Intake...")
  
  # 1. Quantify missingness including institutional sentinel values
  miss_summary <- features_percent_miss(
    raw_df, 
    na_list = c("999", "-99", "Unknown", "Refused", "")
  )
  
  # 2. Flag variables with severe missingness (> 30%)
  flagged_vars <- miss_summary |>
    filter(PctNa > 0.30)
  
  # 3. Screen continuous lab values for extreme clinical outliers
  num_cols <- raw_df |> select(where(is.numeric))
  outliers <- lapply(num_cols, detect_outliers)
  
  list(
    missing_summary = miss_summary,
    severe_missing  = flagged_vars,
    outliers        = outliers
  )
}
```

### Pattern 2: Regulatory missingness tables for Study Protocols & CSRs

Clinical Study Reports (CSRs) submitted to regulatory agencies (FDA, EMA) require
clear tables of missing baseline covariates:

```{r pattern_regulatory, eval = FALSE}
# Export publication-ready missingness tables
csr_missing_table <- features_percent_miss(study_data) |>
  rename(
    `Clinical Variable` = Feature,
    `Missing Count (n)` = N_Missing,
    `Missing Rate (%)`  = PctNa
  ) |>
  mutate(`Missing Rate (%)` = sprintf("%.1f%%", `Missing Rate (%)` * 100))

# Print or write to Excel/RTF
writexl::write_xlsx(csr_missing_table, "tables/CSR_Table_14_1_Missingness.xlsx")
```

### Pattern 3: Outlier screening preserving subject record keys

Flag outliers while maintaining patient identifiers for case review with clinical
investigators:

```{r pattern_outliers, eval = FALSE}
# Screen lab values and join back to patient IDs
iqr_audit <- detect_outliers(study_data$troponin_t)

flagged_patients <- study_data |>
  slice(c(iqr_audit$mild_indices, iqr_audit$extreme_indices)) |>
  select(patient_id, site_id, troponin_t) |>
  mutate(
    flag = ifelse(patient_id %in% study_data$patient_id[iqr_audit$extreme_indices],
                  "EXTREME_OUTLIER", "MILD_OUTLIER")
  )
```

---

## Summary

`TempleCBE` delivers a clean, auditable, and non-destructive exploratory workflow
engineered specifically for the realities of clinical data. By combining
`SumNa()`, `features_percent_miss()`, `missmap()`, `detect_outliers()`, and `add_missing()` with
downstream tuned imputation (`missforest_sweep_mtry()` and `missranger_sweep_mtry()`),
biostatisticians can uphold rigorous data governance standards from initial data
intake through regulatory delivery.


