---
title: "Penalized Cox Models and Nested Cross-Validation for Start/Stop Survival Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{02. Penalized Cox Models and Nested Cross-Validation for Start/Stop Survival Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  dev = "png"
)
required <- c("glmnet", "survival", "rsample", "yardstick", "recipes", "tidymodels")
knitr::opts_chunk$set(eval = all(vapply(required, requireNamespace, logical(1), quietly = TRUE)))
```

## Why start/stop data needs care

In counting-process (start/stop) layout, each subject has one row per
interval, `Surv(tstart, tstop, status)`, so covariates can change over
follow-up. Three things go wrong if those rows are treated as independent
observations:

1. **Resampling leaks.** Row-level folds, like those of `glmnet::cv.glmnet()`,
   put some of a subject's intervals in the analysis set and others in the
   assessment set.
2. **Scoring is wrong, silently.** yardstick's survival metrics expect one
   row per subject with right-censored truth. Given start/stop truth, they
   return numbers without complaint, but those numbers count every interval as
   its own subject.
3. **Censoring is ignored.** Brier scores need inverse-probability-of-censoring
   weights, or subjects lost to follow-up bias the score.

`TempleCBE` handles each of these:

| Function | What it does |
|---|---|
| `coxnet()` | Elastic-net Cox model (glmnet) with formula, recipe, and x/y interfaces (hardhat) |
| `cv_coxnet()` | Chooses `mixture` and `penalty` on subject-grouped folds with a yardstick metric set |
| `nested_cv_coxnet()` | Estimates how well that whole tuning procedure generalizes |
| `surv_subject_truth()`, `censoring_km()`, `graf_weights()`, `add_graf_weights()` | Score any start/stop model with yardstick |

## Comparison with existing R survival & modeling ecosystems

Survival analysis and prognostic modeling in R are supported by several foundational
packages, including [survival](https://cran.r-project.org/package=survival),
[glmnet](https://glmnet.stanford.edu/), [tidymodels](https://www.tidymodels.org/)
([censored](https://censored.tidymodels.org/), [yardstick](https://yardstick.tidymodels.org/)),
and specialized evaluation packages like `pec` and `survAUC`. However, applying
these frameworks to real-world clinical cohorts with longitudinal visits,
time-dependent covariates, and multi-center structures reveals substantial hurdles:

### Existing tools and their limitations in longitudinal clinical modeling

1. **`survival` (Therneau)**: The foundational baseline for semi-parametric Cox
   proportional hazards and Kaplan-Meier estimation. While excellent for classical
   low-dimensional epidemiological inference, it lacks high-dimensional
   regularization (lasso, elastic net), automated cross-validation, and tidy
   resampling integration.
2. **`glmnet::cv.glmnet(family = "cox")` (Friedman et al.)**: The standard
   penalized Cox implementation. However:
   - **Data leakage on start/stop data**: `cv.glmnet` constructs folds by randomly
     partitioning individual *rows*. In counting-process data (`Surv(tstart, tstop, status)`),
     this splits intervals belonging to the *same patient* across training and validation
     folds, leaking patient identity and future biomarker trajectory into the validation set.
   - **No cluster/site-level grouping**: It cannot hold out entire clinical sites
     or hospital clusters to evaluate generalizability across centers.
   - **Suboptimal metric selection**: It chooses the penalty parameter $\lambda$
     using partial likelihood deviance or Harrell's C-index, without support for
     time-dependent calibration, Integrated Brier Scores (IBS), or IPCW adjustments.
   - **No modern formula/recipe preprocessing**: Predictors must be supplied as a
     raw numeric design matrix ($x$), forcing manual one-hot encoding outside the cross-validation loop.
3. **`tidymodels` / `censored` / `yardstick`**: The modern tidy modeling framework.
   While `censored` bridges `parsnip` to survival models, its longitudinal counting-process
   support remains limited:
   - **Silent scoring failures**: `yardstick` survival metrics expect one row per
     subject with right-censored truth (`Surv(time, event)`). When provided with
     counting-process intervals (`Surv(tstart, tstop, event)`), `yardstick` returns
     numbers without error, but silently treats *every interval as an independent patient*,
     biasing both Brier scores and ROC AUCs.
   - **Missing integrated nested CV for penalized Cox**: Evaluating regularized Cox
     models across an $\alpha$-mixture and $\lambda$-penalty grid with subject-level
     IPCW weights requires bespoke orchestration.
4. **`pec` / `riskRegression` / `survAUC`**: Provide inverse-probability-of-censoring
   weights and prediction error curves, but operate as standalone legacy systems
   decoupled from `recipes` preprocessing and the `tidymodels` grammar.

### Where `TempleCBE` provides distinct value

`TempleCBE` unifies high-dimensional regularization, tidy recipes, and rigorous
counting-process survival theory into an integrated, leak-free pipeline:

- **Strict patient- and site-grouped resampling**: `cv_coxnet()` and `nested_cv_coxnet()`
  enforce grouping by patient (`group = "patient_id"`) or site (`group = "site"`),
  guaranteeing that all longitudinal intervals for a patient stay within either the
  analysis set or the assessment set.
- **Formal start/stop to subject-truth translation**: `surv_subject_truth()` collapses
  multi-interval records into the valid single-row-per-subject truth required by
  `yardstick`, verifying that intervals do not overlap and that events occur only
  at terminal visits.
- **Inverse-probability-of-censoring (Graf) weighting**: `censoring_km()`,
  `graf_weights()`, and `add_graf_weights()` estimate the reverse Kaplan-Meier
  censoring distribution on the training set and attach time-dependent IPCW weights
  to test predictions, ensuring Brier scores and ROC curves remain unbiased.
- **True nested cross-validation**: `nested_cv_coxnet()` optimizes hyperparameters
  strictly on inner folds and scores untouched outer assessment folds, delivering
  an honest estimate of clinical generalizability.
- **Mixed continuous and categorical feature spaces**: Through `step_famd()`,
  `TempleCBE` integrates Factor Analysis of Mixed Data directly into `recipes`,
  allowing joint dimensionality reduction across continuous clinical labs and
  discrete clinical stages.
- **Translational laboratory export**: `km_summary_to_prism()` translates survival
  curves into GraphPad Prism-compatible format for oncology and pharmacology deliverables.

### Feature comparison matrix

| Dimension | `survival` | `glmnet::cv.glmnet` | `tidymodels` (native) | `TempleCBE` |
|:---|:---|:---|:---|:---|
| **Counting-process (`tstart, tstop`)** | Full support | Partial (leaks rows in CV) | Limited in `censored` | Full leak-free support |
| **Resampling grouping** | Manual | No (row-level only) | `group_vfold_cv()` | Native patient & site grouping |
| **Penalty selection metric** | None (unpenalized) | Deviance or Concordance | Depends on metric set | Time-dependent IBS & Concordance |
| **IPCW adjustment for start/stop** | Manual | None | Silent row-level miscount | Automated via `graf_weights()` |
| **Tidy recipes integration** | None | None | Standard recipes | Full `recipes` + `step_famd()` |
| **Nested cross-validation engine** | None | None | Manual multi-loop script | Automated via `nested_cv_coxnet()` |
| **Mixed-data dimensionality reduction** | None | None | `step_pca` (numeric only) | `step_famd` (mixed continuous/discrete) |
| **GraphPad Prism deliverable export** | None | None | None | Built-in via `km_summary_to_prism()` |

---

## Specific operational and statistical gaps addressed

### 1. Resampling leakage in longitudinal patient cohorts
In clinical studies tracking chronic disease (e.g. cardiovascular disease, oncology,
sepsis in the ICU), patients undergo repeated visits with drifting biomarkers. When
penalized models are cross-validated at the row level, a model trained on visit 1
and visit 3 is tested on visit 2 of the *same patient*. This produces artificially
deflated prediction error and false optimism. `TempleCBE` guarantees zero leakage
across all cross-validation stages.

### 2. Silent metric bias in counting-process evaluation
Standard evaluation metrics assume each observation corresponds to one independent
individual. Scoring longitudinal intervals directly artificially inflates sample
size (e.g. 150 patients with 4 visits each scored as 600 independent patients) and
ignores informative censoring. `TempleCBE` provides the necessary mathematical
transformations (`surv_subject_truth()` + `graf_weights()`) so that `yardstick`
evaluates patients properly.

### 3. High-dimensional clinical feature reduction for mixed data
Clinical prediction models frequently ingest hundreds of variables combining
continuous laboratory biomarkers with categorical ICD-10 diagnostic codes, cancer
stages, and medication history. Standard PCA cannot handle categorical variables
without arbitrary one-hot expansion that distorts covariance. `TempleCBE::step_famd()`
provides mathematically principled dimensionality reduction for mixed data within
standard `tidymodels` workflows.

---

```{r setup}
library(TempleCBE)
library(tidymodels)
library(survival)
tidymodels_prefer()
```

## Simulated data

Each patient has one to four visits at irregular times, belongs to one of five
sites, and has a marker that drifts between visits. An event can only occur
in a patient's last interval.

```{r generate_data}
set.seed(42)
sim_data <- purrr::map(seq_len(150), function(i) {
  n_visits <- sample(1:4, 1)
  tstop <- round(cumsum(runif(n_visits, 20, 40)))
  age <- rnorm(1, mean = 58, sd = 10)
  treatment <- sample(c("Control", "Treated"), 1)
  marker <- rnorm(1) + cumsum(rnorm(n_visits, sd = 0.3))
  risk <- 0.03 * (age - 58) + 0.8 * marker[n_visits] - 0.6 * (treatment == "Treated")
  tibble(
    patient_id = sprintf("PT_%03d", i),
    site = paste0("site_", i %% 5 + 1),
    tstart = c(0, head(tstop, -1)),
    tstop = tstop,
    status = c(rep(0L, n_visits - 1), rbinom(1, 1, plogis(risk))),
    age = age,
    bmi = rnorm(1, mean = 27, sd = 4),
    marker = marker,
    treatment = factor(treatment, levels = c("Control", "Treated"))
  )
}) |>
  purrr::list_rbind()

sim_data
```

Collapsed to one row per patient, this is the truth yardstick needs:

```{r truth}
surv_subject_truth(Surv(sim_data$tstart, sim_data$tstop, sim_data$status), sim_data$patient_id)
```

## Fitting one model: `coxnet()`

`coxnet()` fits the whole glmnet regularization path. The formula interface
expands factors into indicator columns.

```{r coxnet}
fit <- coxnet(
  Surv(tstart, tstop, status) ~ age + bmi + marker + treatment,
  data = sim_data, mixture = 0.5, penalty = 0.01
)
fit
tidy(fit)
```

Predictions follow tidymodels conventions: `.pred_linear_pred` (larger means
longer survival), or a `.pred` list-column of survival probabilities.

```{r coxnet_predict}
predict(fit, sim_data[1:3, ], type = "survival", eval_time = c(30, 60, 90))$.pred[[1]]
```

## Choosing `mixture` and `penalty`: `cv_coxnet()`

With a recipe, preprocessing is learned inside each fold. The outcome is a
`Surv` column; the patient identifier gets the `"id"` role, which
`cv_coxnet()` uses to group folds and score patients.

```{r cv_coxnet}
sim_data$surv <- Surv(sim_data$tstart, sim_data$tstop, sim_data$status)

cox_rec <- recipe(surv ~ age + bmi + marker + treatment + patient_id + site, data = sim_data) |>
  update_role(patient_id, new_role = "id") |>
  update_role(site, new_role = "site") |>
  step_dummy(treatment) |>
  step_normalize(all_numeric_predictors())

set.seed(2026)
cv <- cv_coxnet(cox_rec, sim_data, mixture = c(0.25, 0.5, 1), v = 5, nlambda = 30)
cv
```

By default every fold is scored with the integrated Brier score (used to
choose), concordance, and the time-specific Brier score and ROC AUC, at deciles
of the observed event times.

```{r cv_metrics}
collect_metrics(cv) |>
  filter(.metric %in% c("brier_survival_integrated", "concordance_survival")) |>
  arrange(.metric, mean)
autoplot(cv)
tidy(cv, penalty = "lambda.1se")
```

Folds can be grouped by a coarser unit than the patient. With
`group = "site"`, each fold holds out whole sites:

```{r cv_sites}
set.seed(2026)
cv_sites <- cv_coxnet(
  cox_rec, sim_data, group = "site", v = 5, nlambda = 20,
  metrics = metric_set(brier_survival_integrated, concordance_survival)
)
cv_sites$best_by_mixture
```

## How well does tuning generalize? `nested_cv_coxnet()`

A cross-validated score at the chosen settings is optimistic: the same
assessment sets chose the settings. Nested cross-validation tunes on inner
resamples, refits on each outer analysis set, and scores the untouched outer
assessment set.

```{r nested}
set.seed(2026)
nested_folds <- nested_cv(
  sim_data,
  outside = group_vfold_cv(group = patient_id, v = 3),
  inside = group_vfold_cv(group = patient_id, v = 3)
)

nested <- nested_cv_coxnet(
  nested_folds, cox_rec,
  mixture = c(0.5, 1), rule = "1se", nlambda = 20,
  metrics = metric_set(brier_survival_integrated, concordance_survival)
)
nested |> select(id, mixture, penalty)
collect_metrics(nested)
```

## Scoring any start/stop model with yardstick

The same helpers score predictions from any model. Here, a `coxnet()` fit
predicts each test patient's survival from their baseline covariates:

```{r any_model}
set.seed(2026)
split <- group_initial_split(sim_data, group = patient_id)
train <- training(split)
test <- testing(split)
times <- c(30, 60, 90)

model <- coxnet(
  Surv(tstart, tstop, status) ~ age + bmi + marker + treatment,
  data = train, mixture = cv$mixture, penalty = cv$lambda_min
)

train_truth <- surv_subject_truth(train$surv, train$patient_id)
test_truth <- surv_subject_truth(test$surv, test$patient_id)
baseline_rows <- test |> filter(tstart == 0) |> arrange(patient_id)

scored <- test_truth |>
  bind_cols(predict(model, baseline_rows, type = "survival", eval_time = times)) |>
  add_graf_weights(censoring = censoring_km(train_truth$.truth))

brier_survival_integrated(scored, truth = .truth, .pred)
```

## Existing analysis code: `glmnet_IBS()`

`glmnet_IBS()`, `tune_over_alpha()`, and `summarize_tune_results()` keep their
arguments. They now use `cv_coxnet()` underneath, so the penalty is chosen on
patient-grouped folds and the IBS uses censoring weights.

```{r glmnet_ibs}
legacy_rec <- recipe(~ age + bmi + marker + treatment + patient_id + tstart + tstop + status, data = sim_data) |>
  update_role(patient_id, tstart, tstop, status, new_role = "id variable") |>
  step_dummy(treatment) |>
  step_normalize(all_numeric_predictors())

set.seed(2026)
glmnet_IBS(
  split, alpha = 0.5, recipe = legacy_rec,
  feature_names = c("age", "bmi", "marker", "treatment_Treated"),
  eval_time = times, id_col = "patient_id", internal_folds = 3
)
```

---

## Highlighting core TempleCBE survival and modeling utilities

Beyond the core `coxnet()` modeling functions, `TempleCBE` includes several
specialized utilities that address frequent friction points in biostatistical
practice:

### 1. Mixed-data dimensionality reduction with `step_famd()`

Clinical prognostic modeling frequently involves datasets with dozens of
continuous lab values (e.g. creatinine, bilirubin, blood pressure) alongside
categorical clinical indicators (e.g. disease stage, smoking status, sex,
treatment center). Standard Principal Component Analysis ([recipes::step_pca()])
only operates on numeric columns; dummy-encoding categorical variables before PCA
distorts the geometric distances because binary indicators have different variances
than continuous variables.

`step_famd()` provides a native `recipes` step for **Factor Analysis of Mixed Data**
(Pagès 2004, implemented via `FactoMineR::FAMD`):
- Continuous variables are normalized to unit variance.
- Categorical variables are dummy-coded and weighted inversely by their category
  frequencies, ensuring each variable contributes proportionally regardless of its
  number of levels.

```{r step_famd_demo, eval = FALSE}
library(recipes)

# Define clinical recipe with mixed numeric and categorical predictors
mixed_rec <- recipe(surv ~ age + bmi + marker + treatment + site + patient_id, data = sim_data) |>
  update_role(patient_id, new_role = "id") |>
  # Apply FAMD to compress continuous and discrete predictors into 3 components
  step_famd(age, bmi, marker, treatment, site, num_comp = 3, prefix = "FAMD_")

# Train and inspect component contributions
prepped_famd <- prep(mixed_rec, training = sim_data)
baked_features <- bake(prepped_famd, new_data = NULL)
head(baked_features)

# Inspect variance explained by components
tidy(prepped_famd, number = 1, type = "variance")
```

`step_famd()` integrates directly into `cv_coxnet()` and `nested_cv_coxnet()`,
ensuring that components are learned strictly inside each training fold without
leakage.

### 2. Framework-agnostic survival scoring with `surv_helpers`

The helper functions in `surv_helpers.R` are not restricted to `coxnet()`: they
enable **any** model (e.g. random survival forests via `ranger`, gradient boosted
survival trees via `xgboost`, or deep learning survival models) to be properly
evaluated on longitudinal start/stop cohorts using `yardstick`:

1. `surv_subject_truth()`: Collapses start/stop interval data to a single row
   per patient with right-censored truth.
2. `censoring_km()`: Estimates the reverse Kaplan-Meier censoring distribution
   $\hat{G}(t) = P(C > t)$ on the training cohort.
3. `graf_weights()` / `add_graf_weights()`: Adds Inverse Probability of Censoring
   Weights (Graf et al. 1999) to prediction tibbles.

```{r external_model_scoring, eval = FALSE}
# Scoring any external model predictions on test subjects:
test_truth <- surv_subject_truth(test_longitudinal$surv, test_longitudinal$patient_id)
train_censoring <- censoring_km(train_subject_truth$.truth)

# Combine test truth with external model predictions and add IPCW weights
scored_predictions <- test_truth |>
  bind_cols(external_model_predictions) |>
  add_graf_weights(censoring = train_censoring)

# Compute proper, unbiased integrated Brier score
yardstick::brier_survival_integrated(scored_predictions, truth = .truth, .pred)
```

### 3. Distributed elastic net tuning with `survival_metrics`

When searching over a broad grid of elastic net mixing parameters ($\alpha$) or
evaluating competing candidate clinical formulas, `tune_over_alpha()` and
`summarize_tune_results()` orchestrate parallel evaluation using `furrr`:

```{r tune_alpha_demo, eval = FALSE}
library(furrr)
plan(multisession, workers = 4)

# Parallel grid search across alpha values for an rsample split
alpha_results <- tune_over_alpha(
  split,
  recipe = cox_rec,
  num_alpha_values = 10,
  internal_folds = 5,
  progress = TRUE
)

# Summarize results across all inner splits of a nested resample
nested_alpha_summary <- summarize_tune_results(nested_folds$inner_resamples[[1]])
```

### 4. Translating survival curves to GraphPad Prism with `km_summary_to_prism()`

In academic medical centers, biostatisticians frequently collaborate with
laboratory scientists, pharmacologists, and clinical trialists who require
Kaplan-Meier survival data formatted for **GraphPad Prism** for manuscript figures
and FDA briefing documents.

Prism requires a unique table format: an `X` column of timepoints, with separate
columns per experimental group holding `1` for events, `0` for censored patients,
and `NA` for other groups. `km_summary_to_prism()` converts standard
`survival::survfit` summary tables directly into this layout:

```{r prism_demo, eval = FALSE}
# Fit standard stratified Kaplan-Meier curve
fit_km <- survfit(Surv(time, status) ~ treatment, data = lung)
km_summary <- summary(fit_km, censored = TRUE)

km_df <- data.frame(
  time     = km_summary$time,
  strata   = km_summary$strata,
  n.risk   = km_summary$n.risk,
  n.event  = km_summary$n.event,
  n.censor = km_summary$n.censor
)

# Convert to GraphPad Prism layout and export to Excel
prism_table <- km_summary_to_prism(
  km_df,
  out_xlsx = "deliverables/Prism_Survival_Table.xlsx"
)
head(prism_table)
```

---

## How others can use this technology (Adoption Patterns)

Here are five practical adoption patterns for biostatistical and clinical data
science teams:

### Pattern 1: Longitudinal EHR survival modeling with patient grouping
When modeling electronic health records with repeated clinical visits, replace
standard row-level splitting with patient-grouped folds:

```{r pattern_ehr, eval = FALSE}
# Ensure patient-grouped folds
cv_fit <- cv_coxnet(
  recipe = clinical_rec,
  data = ehr_longitudinal_data,
  group = "patient_id",
  mixture = 0.5,
  v = 5
)
```

### Pattern 2: Mixed clinical & biomarker dimensionality reduction
Combine high-dimensional lab panels and discrete clinical stages using `step_famd()`
inside a `tidymodels` workflow:

```{r pattern_famd, eval = FALSE}
rec <- recipe(surv ~ ., data = cohort) |>
  update_role(patient_id, new_role = "id") |>
  step_famd(all_predictors(), num_comp = 5)

fit <- coxnet(rec, data = cohort, mixture = 1.0, penalty = 0.05)
```

### Pattern 3: Multi-center clinical trial generalizability assessment
Evaluate whether a prognostic signature generalises across hospital sites by
grouping cross-validation folds at the institutional site level:

```{r pattern_multicenter, eval = FALSE}
# Hold out entire clinical sites during validation
site_cv <- cv_coxnet(
  recipe = clinical_rec,
  data = multi_site_study,
  group = "hospital_site",
  v = 10,
  metrics = metric_set(brier_survival_integrated, concordance_survival)
)
```

### Pattern 4: Honest generalization auditing with nested cross-validation
Before presenting prognostic scores to clinical stakeholders or submitting
models for regulatory evaluation, estimate true out-of-sample generalization
using `nested_cv_coxnet()`:

```{r pattern_nested, eval = FALSE}
nested_plan <- nested_cv(
  cohort_data,
  outside = group_vfold_cv(group = patient_id, v = 5),
  inside  = group_vfold_cv(group = patient_id, v = 5)
)

honest_metrics <- nested_cv_coxnet(nested_plan, clinical_rec, mixture = c(0.5, 1.0))
collect_metrics(honest_metrics)
```

### Pattern 5: Translational laboratory handoff
Seamlessly bridge biostatistical analysis in R with laboratory GraphPad Prism
figure creation using `km_summary_to_prism()`:

```{r pattern_prism, eval = FALSE}
# Deliver Prism-ready survival spreadsheets to clinical collaborators
km_summary_to_prism(
  km_table,
  out_xlsx = "reports/figures/Figure1_KaplanMeier_Prism.xlsx"
)
```

---

## Summary

`TempleCBE` fills the critical gap between classical low-dimensional survival
analysis and modern high-dimensional machine learning on longitudinal clinical
cohorts. By pairing leak-free grouped resampling (`cv_coxnet()`, `nested_cv_coxnet()`)
with proper IPCW scoring (`surv_helpers`), mixed-data feature reduction (`step_famd()`),
and translational handoff tools (`km_summary_to_prism()`), it provides a robust,
reproducible foundation for clinical survival modeling.

