---
title: "Exploratory Data Analysis, Missingness Auditing, and Normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Exploratory Data Analysis, Missingness Auditing, and Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dev = "png"
)
```

## Overview

Clinical datasets and electronic health records (EHR) frequently exhibit complex missingness patterns, non-standard missing codes (e.g. `"999"`, `"NA"`, `""`), and extreme statistical outliers. The **`TempleCBE`** package provides domain-agnostic tools to streamline exploratory data analysis (EDA), quantify missingness, normalize numerical features, and flag outliers cleanly.

```r
library(TempleCBE)
library(dplyr)
```

---

## 1. Auditing Missing Data

### Total Missingness with `SumNa()`

`SumNa()` counts missing values in vectors, data frames, or matrices, supporting explicit missing value codes (such as `"NA"` strings or empty strings):

```{r sum_na_demo}
library(TempleCBE)

# Sample vector with default and non-standard missing values
vec <- c(12, 14, NA, 999, 18, "")

# Count standard NAs
SumNa(vec)

# Count NAs including custom missing codes
SumNa(vec, na_list = c("999", ""))
```

### Feature-Level Missingness Table

`features_percent_miss()` generates a detailed summary table sorted descending by missingness percentage:

```{r features_percent_miss_demo}
# Synthetic clinical dataframe
df_clinical <- data.frame(
  patient_id = 1:6,
  age = c(45, 52, NA, 61, 38, NA),
  systolic_bp = c(120, 135, NA, 140, 118, 125),
  bmi = c(22.5, NA, 29.1, NA, 31.0, NA),
  smoker = c("No", "Yes", "", "No", "NA", "Yes"),
  stringsAsFactors = FALSE
)

# Audit missingness with custom codes
features_percent_miss(df_clinical, na_list = c("", "NA"))
```

### Missingness Visualizations

Visualize missingness across all features using `plot_features_percent_miss()` or S3 `plot()`:

```r
# Plot feature missingness chart
plot_features_percent_miss(df_clinical)

# Plot top 3 missing features
plot_features_percent_miss(df_clinical, top_n = 3)
```

Or view complete missingness heatmaps with `missmap()`:

```r
# Render missingness heatmap
missmap(df_clinical)
```

---

## 2. Feature Normalization

`TempleCBE` includes three standard feature scaling transformations:

- **`min_max_norm()`**: Scales values to the range $[0, 1]$.
- **`z_norm()`**: Standardizes features to mean $0$ and standard deviation $1$.
- **`range_norm()`**: Scales relative to combined distribution bounds across columns.

```{r norm_demo}
x <- c(10, 20, 30, 40, 50)

# Min-Max Normalization
min_max_norm(x)

# Z-Score Normalization
z_norm(x)
```

---

## 3. Outlier Detection

`detect_outliers()` evaluates numeric vectors or data frame columns using Interquartile Range (IQR) thresholding, distinguishing between **MILD** (1.5 × IQR) and **EXTREME** (3.0 × IQR) outliers.

```{r outlier_demo}
data_vals <- c(12, 14, 15, 14, 16, 15, 14, 45, 120)

# Detect mild and extreme outliers
detect_outliers(data_vals)
```

---

## Summary

Combining `SumNa()`, `features_percent_miss()`, `min_max_norm()`, and `detect_outliers()` provides a reproducible, auditable workflow for preparing clinical data for downstream statistical modeling.
