---
title: "01. Exploratory Data Analysis, Missingness Auditing, and Normalization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{01. Exploratory Data Analysis, Missingness Auditing, and Normalization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  dev = "png"
)
```

## Overview

In clinical trials, electronic health records (EHR), and epidemiological
registries, data cleanliness is rarely a given. Datasets frequently exhibit
complex missingness patterns, non-standard missing codes (e.g. `"999"`, `"-99"`,
`"Unknown"`, `"Refused"`, `""`), and extreme statistical outliers from data-entry
errors or severe disease phenotypes.

Standard data science tools often assume missingness is cleanly represented as
`NA`, requiring destructive multi-pass preprocessing before exploratory analysis
can even begin. The **`TempleCBE`** package provides domain-tailored biostatistical
tools to audit non-standard missingness non-destructively, rank feature completeness,
standardize numerical features, classify clinical outliers under formal IQR
standards, and transition seamlessly into tuned machine-learning imputation.

```{r setup}
library(TempleCBE)
library(dplyr)
library(ggplot2)
```

---

## Comparison with existing R EDA & missingness tools

The R ecosystem offers several packages for exploratory data analysis and
missingness visualization, notably [naniar](https://naniar.njtierney.com/),
[visdat](https://docs.ropensci.org/visdat/), [mice](https://amices.org/mice/),
and general EDA toolkits like `DataExplorer` and `summarytools`. Understanding
where `TempleCBE` fits alongside these frameworks clarifies its value for
clinical biostatisticians and data managers.

### Existing tools and their limitations in clinical workflows

- **`naniar` & `visdat`**: Pioneered tidy representation and visualization of
  missing data (`vis_miss()`, `gg_miss_var()`). However, they operate exclusively
  on existing `NA` values. When raw clinical data contains sentinel values (such
  as `"999"`, `"-99"`, or blank strings), users must first apply mutating
  functions like `replace_with_na_all()`, which can accidentally alter legitimate
  numerical values, modify factor levels, or mutate data types.
- **`mice` / `Amelia`**: Primarily multiple imputation engines. While they
  include basic diagnostics (e.g. `md.pattern()`), their tools are oriented
  toward parameter checking rather than automated feature-level reporting for
  regulatory deliverables.
- **Base R / `dplyr` summaries**: Simple calls like `colSums(is.na(df))` provide
  quick counts, but lack built-in percentage calculations, descending order
  sorting, sentinel code awareness, and publication-ready formatting.

### Where `TempleCBE` provides distinct value

`TempleCBE` was built specifically for the constraints of institutional
biostatistics core facilities:

1. **Zero-mutation non-standard code auditing**: `SumNa()` and
   `features_percent_miss()` natively accept vectors of custom missing codes
   (`na_list = c("999", "-99", "", "Unknown", "NA")`). They compute exact
   missingness counts and percentages *without altering or coercing the underlying
   data frame*.
2. **Two-tier clinical outlier governance**: While generic outlier functions often
   rely on simple z-score cutoffs or arbitrary quantiles, `detect_outliers()`
   implements Tukey's formal biostatistical standard, explicitly distinguishing
   between **MILD** ($1.5 \times \text{IQR}$) and **EXTREME** ($3.0 \times \text{IQR}$)
   outliers and returning structured classification flags that preserve patient keys.
3. **Controlled imputation benchmarking**: The `add_missing()` function allows
   statisticians to simulate controlled Missing Completely at Random (MCAR)
   mechanisms on complete datasets, tracking the exact row-and-column coordinates
   of every amputed cell to empirically validate downstream imputation performance.
4. **Seamless transition to tuned imputation**: `TempleCBE` links missingness
   auditing directly to high-performance imputation engines (`missforest_sweep_mtry()`
   and `missranger_sweep_mtry()`) that automatically optimize random forest hyperparameters
   (`mtry`) independently for each clinical feature.

### Feature comparison matrix

| Dimension | Base / Tidyverse | `naniar` / `visdat` | `TempleCBE` |
|:---|:---|:---|:---|
| **Multi-code missingness** | Manual recoding required | Requires mutating `replace_with_na()` | Built-in via `na_list` argument (zero mutation) |
| **Feature-level ranking** | Multi-line `summarise()` | `miss_var_summary()` | `features_percent_miss()` (sorted tibble + S3 plot) |
| **Missingness heatmap** | Not built-in | `vis_miss()` | `missmap()` (custom clinical palette & grouping) |
| **Outlier classification** | Manual IQR calculations | Not covered | Two-tier Tukey standard (MILD: 1.5×, EXTREME: 3.0×) |
| **Feature normalization** | `scale()` (matrix output) | Not covered | `min_max_norm()`, `z_norm()`, `range_norm()` |
| **Controlled amputation** | External (`mice::ampute`) | Limited | `add_missing()` with tracked cell coordinates |
| **Imputation handoff** | External | External | Integrated `missforest_sweep_mtry()` / `missranger_sweep_mtry()` |

---

## 1. Starting with a Full Synthetic Clinical Cohort

To demonstrate missingness auditing and empirically evaluate imputation accuracy,
we begin with a relatively full synthetic cohort of $N = 120$ clinical patients.
In real-world studies, true underlying values are unknown once lost. By generating
a complete cohort with realistic physiological correlations first, we establish a
verifiable **ground truth** that enables downstream recovery benchmarking:

```{r generate_cohort}
set.seed(2026)
n <- 120

# Realistic clinical covariance structure
age <- round(rnorm(n, mean = 58, sd = 9))
bmi <- round(rnorm(n, mean = 27.0, sd = 3.8), 1)
systolic_bp <- round(95 + 0.45 * age + 0.65 * bmi + rnorm(n, mean = 0, sd = 7))
cholesterol <- round(155 + 0.55 * age + 1.10 * bmi + rnorm(n, mean = 0, sd = 15))
glucose <- round(72 + 0.35 * age + 0.95 * bmi + rnorm(n, mean = 0, sd = 10))
smoker <- factor(sample(c("No", "Yes"), n, replace = TRUE, prob = c(0.65, 0.35)))
hypertension <- factor(ifelse(systolic_bp >= 135 | (age >= 62 & bmi >= 28), "Yes", "No"))
disease_stage <- factor(sample(c("Stage I", "Stage II", "Stage III", "Stage IV"), n,
  replace = TRUE, prob = c(0.30, 0.30, 0.25, 0.15)
))

cohort_clean <- tibble::tibble(
  patient_id    = sprintf("PT_%03d", seq_len(n)),
  age           = age,
  systolic_bp   = systolic_bp,
  bmi           = bmi,
  cholesterol   = cholesterol,
  glucose       = glucose,
  smoker        = smoker,
  hypertension  = hypertension,
  disease_stage = disease_stage
)

# Preview the clean cohort
head(cohort_clean, 6)

# Verify complete data integrity (zero missingness)
SumNa(cohort_clean)
```

The dataset contains:
- Continuous biomarkers: `age`, `systolic_bp`, `bmi`, `cholesterol`, and `glucose`.
- Categorical clinical factors: `smoker`, `hypertension`, and `disease_stage`.
- Unique identifier: `patient_id`.

---

## 2. Controlled Amputation with `add_missing()`

In validation studies, biostatisticians simulate controlled Missing Completely at Random
(MCAR) mechanisms using `add_missing()`. This allows us to observe how data-auditing
tools detect missingness and how machine-learning algorithms reconstruct true values.

Here, we amputate **15% missingness** across 7 clinical predictors, leaving `patient_id`
and `disease_stage` 100% complete:

```{r ampute_cohort}
set.seed(42)
amputed <- add_missing(
  cohort_clean,
  cols = c(age, systolic_bp, bmi, cholesterol, glucose, smoker, hypertension),
  pct_na = 0.15
)

# Inspect the amputed cohort preview
head(amputed, 6)
```

### Tracking masked coordinates for ground-truth validation

A crucial feature of `add_missing()` is that it attaches an attribute named
`"missing_cells"`. This attribute stores a tibble of every masked cell's exact `row`
number and `feature` name:

```{r missing_cells_attr}
# Extract the ground-truth coordinate lookup table
missing_cells <- attr(amputed, "missing_cells")

cat("Total cells amputed:", nrow(missing_cells), "\n")
head(missing_cells, 8)
```

Because most `dplyr` verbs strip custom attributes upon transformation, always extract
`missing_cells <- attr(amputed, "missing_cells")` immediately following amputation
when benchmarking imputation pipelines.

### Heterogeneous missingness: Column-specific proportions via vector `pct_na`

In real-world clinical cohorts, different variables rarely share an identical
missingness rate. Routine clinical covariates (such as `age` or `bmi`) may only be
missing in 5--10% of records, whereas specialized laboratory panels (such as
`cholesterol` or `glucose`) may be uncollected in 25--35% of patients.

`add_missing()` supports a numeric vector for `pct_na` whose length matches the
number of selected columns:

```{r ampute_vector_demo}
# Generate column-specific missingness proportions with runif()
set.seed(42)
cols_ampute <- c("age", "systolic_bp", "bmi", "cholesterol", "glucose", "smoker", "hypertension")

p_vector <- cols_ampute |>
  length() |>
  runif(min = 0.05, max = 0.30) |>
  round(2)

names(p_vector) <- cols_ampute
p_vector

# Amputate each column according to its specific proportion
amputed_hetero <- add_missing(
  cohort_clean,
  cols = c(age, systolic_bp, bmi, cholesterol, glucose, smoker, hypertension),
  pct_na = p_vector
)

# Inspect how features_percent_miss captures the heterogeneous rates
features_percent_miss(amputed_hetero)
```

Notice how `features_percent_miss()` directly reflects each column's distinct
amputation share, and the `missing_cells` attribute accurately tracks the varying
number of masked cells per feature.

---

## 3. Auditing Missing Data

Now that missingness has been induced, we audit the dataset using `TempleCBE`'s
non-destructive auditing suite.

### Total missingness with `SumNa()`

`SumNa()` evaluates total missing cells across an entire vector, data frame, or
matrix. When working with raw hospital extracts containing non-standard placeholders,
specify custom sentinel codes in `na_list`:

```{r sum_na_demo}
# Total missing cells in our amputed cohort (18 rows * 7 features = 126 cells)
SumNa(amputed)

# Demonstrating multi-sentinel code auditing without modifying data types
vec_raw <- c(12, 14, NA, 999, 18, "", "Unknown", -99)
SumNa(vec_raw, na_list = c("999", "-99", "", "Unknown"))
```

### Feature-level completeness table with `features_percent_miss()`

In clinical study preparation, biostatisticians need a sorted summary showing
which variables exceed missingness thresholds (e.g. variables with $>20\%$
missingness that cannot be reliably imputed):

```{r features_percent_miss_demo}
miss_summary <- features_percent_miss(amputed)
miss_summary
```

Notice that:
- All 7 amputed features exhibit exactly `PctNa = 0.15` (18 missing cells out of 120 rows).
- The un-amputed columns (`patient_id` and `disease_stage`) correctly report `PctNa = 0.00`.
- The output is automatically sorted in descending order of missingness proportion.

### Visualizing missingness with `plot_features_percent_miss()`

`plot_features_percent_miss()` (or calling `plot()` on a `features_percent_miss` object)
produces a clean horizontal bar chart displaying feature-level missingness:

```{r plot_miss_summary}
plot_features_percent_miss(amputed)
```

We can also isolate the top features or enforce custom thresholds:

```{r plot_miss_top}
plot_features_percent_miss(amputed, top_n = 5)
```

### Missingness heatmaps with `missmap()`

The `missmap()` function visualizes missingness patterns across the cohort.

#### 1. Patient-level heatmap (`row_order = FALSE`)

Visualizes every patient row and feature column. Rows and columns are sorted by
missingness density, allowing visual identification of patients with clustered
missing labs:

```{r missmap_patient}
missmap(amputed)
```

#### 2. Group-aggregated heatmap (`by_column = ...`)

In clinical trials, auditing missingness aggregated by study arm, hospital site,
or disease stage is critical for trial monitoring:

```{r missmap_stage}
# Aggregate missingness counts per disease stage
missmap(amputed, by_column = disease_stage, fill = "count")
```

---

## 4. Clinical Outlier Governance and Feature Normalization

Before proceeding to imputation, exploratory clinical workflows require screening
for distributional anomalies and scaling features.

### Two-tier Tukey outlier detection

`detect_outliers()` audits numerical features using Tukey's formal IQR criteria:
- **Mild Outliers**: Values lying outside $[Q_1 - 1.5 \times \text{IQR}, Q_3 + 1.5 \times \text{IQR}]$.
- **Extreme Outliers**: Values lying outside $[Q_1 - 3.0 \times \text{IQR}, Q_3 + 3.0 \times \text{IQR}]$.

```{r outlier_demo}
# Screen cholesterol for mild and extreme clinical anomalies
outlier_chol <- detect_outliers(cohort_clean$cholesterol)
outlier_chol
```

### Distributional testing

Before applying parametric survival or regression models, test distributional assumptions
with `distribution_test()`:

```{r dist_test_demo}
distribution_test(cohort_clean$glucose)
```

### Feature normalization

`TempleCBE` provides three standard vector transformations for preparing predictors
for penalized regressions or distance-based algorithms:
- `min_max_norm()`: Scales values strictly to $[0, 1]$.
- `z_norm()`: Centers to mean $\mu = 0$ and scales to $\sigma = 1$.
- `range_norm()`: Scales relative to combined distribution bounds.

```{r norm_demo}
# Sample values from baseline systolic blood pressure
bp_sample <- cohort_clean$systolic_bp[1:5]

tibble::tibble(
  raw     = bp_sample,
  min_max = round(min_max_norm(bp_sample), 3),
  z_score = round(z_norm(bp_sample), 3)
)
```

### Pairwise correlation screening with `corr_test_all()`

Screening for multicollinearity across clinical biomarkers helps identify redundant
covariates before regression modeling and confirms the underlying covariance that
imputation engines exploit:

```{r corr_test_demo}
# Pairwise correlation across continuous clinical biomarkers
corr_tbl <- corr_test_all(
  cohort_clean[, c("age", "systolic_bp", "bmi", "cholesterol", "glucose")]
)
head(corr_tbl, 6)
```

---

## 5. Tuned Random Forest Imputation with `missforest_sweep_mtry()`

Once missingness patterns and outliers have been audited, biostatisticians must
handle missing values. Random forest imputation is widely considered the gold standard
for mixed clinical data because it captures non-linear physiological relationships
and complex interactions without requiring parametric distribution assumptions.

### The `mtry` dilemma in clinical data

Random forest imputation iterates through each variable with missing values,
fitting a forest predicting that variable from all other variables. The single
most impactful hyperparameter is **`mtry`** — the number of candidate variables
sampled at each split:
- If `mtry` is too low, strong predictive biomarkers may be missed at key splits.
- If `mtry` is too high, trees become correlated and overfit noisy clinical features.

Critically, **different clinical variables perform best at different `mtry` values**.
A continuous biomarker (e.g. `systolic_bp`) may achieve minimal error at $mtry = 3$,
whereas a binary factor (e.g. `smoker`) may perform best at $mtry = 1$. Standard
implementations force analysts to choose a single global `mtry`.

### `TempleCBE`'s solution: Independent per-column best `mtry` assembly

`missforest_sweep_mtry()` sweeps across candidate `mtry` values, tracks out-of-bag (OOB)
error for every column at every `mtry`, and pulls each column independently from the
specific run where that column achieved its minimal error.

```{r missforest_run, eval = requireNamespace("missForest", quietly = TRUE)}
# Run missForest sweep across mtry = 1:4
# We exclude patient_id and disease_stage from imputation predictors
res_forest <- missforest_sweep_mtry(
  data            = amputed,
  exclude         = c("patient_id", "disease_stage"),
  mtry_values     = 1:4,
  ntree           = 60,
  maxiter         = 5,
  parallel        = FALSE,
  seed            = 123
)
```

### Inspecting the winning `mtry` per feature

```{r missforest_best, eval = requireNamespace("missForest", quietly = TRUE)}
# Winning mtry and minimum OOB error per column
res_forest$best
```

Notice that:
- Features optimize at **different** `mtry` values (e.g. `smoker` at $mtry = 1$, `hypertension` and `glucose` at $mtry = 2$, `systolic_bp` and `cholesterol` at $mtry = 3$, `age` and `bmi` at $mtry = 4$).
- Categorical features report Proportion of Falsely Classified (**PFC**), while continuous features report Mean Squared Error (**MSE**).
- `imp_data` contains the completed cohort with `patient_id` and `disease_stage` re-attached in place:

```{r missforest_preview, eval = requireNamespace("missForest", quietly = TRUE)}
head(res_forest$imp_data, 6)

# Verify that all missing values have been successfully resolved
SumNa(res_forest$imp_data)
```

---

## 6. Comparing Differences in Errors

A rigorous validation workflow evaluates errors along two dimensions:
1. **Model-reported Out-of-Bag (OOB) errors** across candidate `mtry` values.
2. **Empirical recovery errors** comparing imputed values directly against the known ground truth.

### Comparison 1: Differences in OOB errors across `mtry` values

We first examine how out-of-bag error varies across the swept `mtry` values for each feature:

```{r oob_comparison_table, eval = requireNamespace("missForest", quietly = TRUE)}
# Tidy OOB error across all mtry values
res_forest$oob_error |>
  dplyr::arrange(column, mtry)
```

We can visualize this trajectory to see why variable-wise selection outperforms
a single global `mtry`:

```{r oob_comparison_plot, eval = requireNamespace("missForest", quietly = TRUE), fig.height = 5}
res_forest$oob_error |>
  ggplot(aes(x = mtry, y = error, color = column, group = column)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  facet_wrap(~error_type, scales = "free_y") +
  theme_minimal(base_size = 12) +
  labs(
    title = "Out-of-Bag (OOB) Imputation Error Across mtry Settings",
    subtitle = "Continuous features report MSE; Categorical features report PFC",
    x = "Candidate mtry Value",
    y = "Out-of-Bag Error",
    color = "Clinical Feature"
  ) +
  scale_x_continuous(breaks = 1:4)
```

Notice the key biostatistical insights:
- If a global setting of $mtry = 1$ had been enforced, continuous features like `systolic_bp` and `cholesterol` would suffer significantly elevated MSE.
- If a global setting of $mtry = 4$ had been enforced, binary features like `smoker` and `hypertension` would suffer increased misclassification.
- `TempleCBE`'s independent column assembly selects the global minimum for every variable simultaneously.

### Comparison 2: Imputation Recovery vs. Known Ground Truth

Because we started with a complete dataset (`cohort_clean`) and used `add_missing()`
to induce missingness, we can cross-reference every imputed cell with its original,
unmasked ground truth:

```{r truth_vs_imputed_extract, eval = requireNamespace("missForest", quietly = TRUE)}
# Match imputed cells to true values via missing_cells coordinates
truth_vs_imp <- missing_cells |>
  dplyr::rowwise() |>
  dplyr::mutate(
    true_val    = as.character(cohort_clean[[feature]][row]),
    imputed_val = as.character(res_forest$imp_data[[feature]][row])
  ) |>
  dplyr::ungroup()

# Sample of ground truth vs imputed predictions
head(truth_vs_imp, 10)
```

#### Continuous feature recovery metrics

For continuous biomarkers, we quantify recovery fidelity using:
- **Mean Absolute Error (MAE)**: Average magnitude of prediction errors.
- **Root Mean Squared Error (RMSE)**: Penalizes larger deviations.
- **Pearson Correlation ($r$)**: Measures linear agreement between true and recovered values.

```{r continuous_recovery_metrics, eval = requireNamespace("missForest", quietly = TRUE)}
cont_features <- c("age", "systolic_bp", "bmi", "cholesterol", "glucose")

cont_metrics <- truth_vs_imp |>
  dplyr::filter(feature %in% cont_features) |>
  dplyr::mutate(
    true_num = as.numeric(true_val),
    imp_num  = as.numeric(imputed_val),
    diff     = imp_num - true_num
  ) |>
  dplyr::group_by(feature) |>
  dplyr::summarise(
    n_masked  = dplyr::n(),
    MAE       = round(mean(abs(diff)), 2),
    RMSE      = round(sqrt(mean(diff^2)), 2),
    Corr      = round(cor(imp_num, true_num), 3),
    .groups   = "drop"
  )

cont_metrics
```

#### Categorical feature recovery metrics

For binary clinical factors, we evaluate classification accuracy and misclassification rate:

```{r categorical_recovery_metrics, eval = requireNamespace("missForest", quietly = TRUE)}
cat_features <- c("smoker", "hypertension")

cat_metrics <- truth_vs_imp |>
  dplyr::filter(feature %in% cat_features) |>
  dplyr::group_by(feature) |>
  dplyr::summarise(
    n_masked      = dplyr::n(),
    accuracy      = round(mean(true_val == imputed_val), 3),
    mismatch_rate = round(mean(true_val != imputed_val), 3),
    .groups       = "drop"
  )

cat_metrics
```

### Visualizing recovery fidelity

We can visually evaluate recovery by plotting true values versus imputed values for continuous
predictors:

```{r recovery_scatter, eval = requireNamespace("missForest", quietly = TRUE), fig.height = 4.5}
truth_vs_imp |>
  dplyr::filter(feature %in% c("systolic_bp", "age", "glucose")) |>
  dplyr::mutate(
    true_num = as.numeric(true_val),
    imp_num  = as.numeric(imputed_val)
  ) |>
  ggplot(aes(x = true_num, y = imp_num, color = feature)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey50") +
  geom_point(size = 2.5, alpha = 0.8) +
  facet_wrap(~feature, scales = "free") +
  theme_minimal(base_size = 11) +
  labs(
    title = "True vs. Imputed Values for Amputed Clinical Biomarkers",
    subtitle = "Dashed line indicates perfect recovery (y = x)",
    x = "True Value (Ground Truth)",
    y = "Imputed Value (TempleCBE missForest)",
    color = "Feature"
  )
```

The recovery results confirm that:
- Strongly correlated features like `systolic_bp` and `hypertension` achieve high recovery precision (correlation $r > 0.63$ and classification accuracy $> 94\%$).
- Out-of-bag error tracks closely with true empirical recovery error, confirming that optimizing `mtry` per feature leads to tangible gains in reconstructing real patient data.

---

## 7. Fast, Large-Scale Imputation with `missranger_sweep_mtry()`

For large clinical cohorts (e.g. $> 10,000$ patients or high-dimensional panels),
`missForest` can be computationally intensive. `missranger_sweep_mtry()` leverages
the C++ `ranger` engine and supports Predictive Mean Matching (PMM):

- **Predictive Mean Matching (`pmm.k`)**: Setting `pmm.k > 0` (e.g. `pmm.k = 3`)
  ensures that imputed values are drawn from actual observed patient donor values
  rather than synthetic regression averages — critical for bounded clinical scores
  (e.g. integer counts, survey scores).
- **Parallel evaluation**: Uses `furrr` to evaluate candidate `mtry` values across
  CPU cores in parallel without thread oversubscription.

```{r missranger_sweep_demo, eval = FALSE}
library(furrr)
plan(multisession, workers = 2)

# Fast sweep with predictive mean matching
res_ranger <- missranger_sweep_mtry(
  data        = amputed,
  exclude     = c("patient_id", "disease_stage"),
  mtry_values = 1:4,
  num.trees   = 200,
  pmm.k       = 3,
  seed        = 2026,
  parallel    = TRUE
)

# Inspect winning mtry values from ranger
res_ranger$best
```

### Safety and data governance guardrails

Both sweep functions incorporate essential clinical data safeguards:
1. **Identifier and timestamp protection (`exclude`)**: Clinical metadata (patient
   IDs, specimen barcodes, visit dates) are quarantined from the forest training
   matrix and re-attached unaltered to the final output.
2. **Preventing data fabrication (`max_pct_missing`)**: Features that are $>50\%$
   missing (or any custom threshold) are automatically held out from imputation,
   carried through unimputed, and listed in `excluded_high_missing`.
3. **Refusing all-`NA` columns (`check_no_all_na_columns`)**: Standard `missForest`
   silently drops all-`NA` columns, while `missRanger` silently returns them
   still `NA`. `TempleCBE` explicitly halts and warns the user so schema integrity
   is never compromised.

---

## 8. Practical Adoption Patterns

Here are three common patterns for applying `TempleCBE`'s EDA tools to real-world
studies:

### Pattern 1: Clinical trial data intake audit

Create a standardized data quality intake report for newly exported registry data:

```{r pattern_intake, eval = FALSE}
audit_clinical_intake <- function(raw_df) {
  message("Auditing Clinical Intake...")

  # 1. Quantify missingness including institutional sentinel values
  miss_summary <- features_percent_miss(
    raw_df,
    na_list = c("999", "-99", "Unknown", "Refused", "")
  )

  # 2. Flag variables with severe missingness (> 30%)
  flagged_vars <- miss_summary |>
    filter(PctNa > 0.30)

  # 3. Screen continuous lab values for extreme clinical outliers
  num_cols <- raw_df |> select(where(is.numeric))
  outliers <- lapply(num_cols, detect_outliers)

  list(
    missing_summary = miss_summary,
    severe_missing  = flagged_vars,
    outliers        = outliers
  )
}
```

### Pattern 2: Regulatory missingness tables for Study Protocols & CSRs

Clinical Study Reports (CSRs) submitted to regulatory agencies (FDA, EMA) require
clear tables of missing baseline covariates:

```{r pattern_regulatory, eval = FALSE}
# Export publication-ready missingness tables
csr_missing_table <- features_percent_miss(amputed) |>
  rename(
    `Clinical Variable` = feature,
    `Missing Count (n)` = SumNa,
    `Missing Rate (%)`  = PctNa
  ) |>
  mutate(`Missing Rate (%)` = sprintf("%.1f%%", `Missing Rate (%)` * 100))

# Print or write to Excel/RTF
writexl::write_xlsx(csr_missing_table, "CSR_Table_14_1_Missingness.xlsx")
```

### Pattern 3: Outlier screening preserving subject record keys

Flag outliers while maintaining patient identifiers for case review with clinical
investigators:

```{r pattern_outliers, eval = FALSE}
# Screen lab values and join back to patient IDs
iqr_audit <- detect_outliers(cohort_clean$cholesterol)

flagged_patients <- cohort_clean |>
  slice(c(iqr_audit$mild_indices, iqr_audit$extreme_indices)) |>
  select(patient_id, disease_stage, cholesterol) |>
  mutate(
    flag = ifelse(patient_id %in% cohort_clean$patient_id[iqr_audit$extreme_indices],
      "EXTREME_OUTLIER", "MILD_OUTLIER"
    )
  )
```

---

## Summary

`TempleCBE` delivers a clean, auditable, and non-destructive exploratory workflow
engineered specifically for the realities of clinical data. By combining
`SumNa()`, `features_percent_miss()`, `missmap()`, `detect_outliers()`, and `add_missing()` with
downstream tuned imputation (`missforest_sweep_mtry()` and `missranger_sweep_mtry()`),
biostatisticians can uphold rigorous data governance standards from initial data
intake through regulatory delivery.
