---
title: "Nested Cross-Validation for Longitudinal Survival Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Nested Cross-Validation for Longitudinal Survival Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  dev = "png"
)
```

## Introduction & Overview

In clinical trial data science and observational cohort modeling, evaluating survival outcomes (e.g. time-to-event data with censoring and non-adherence) requires rigorous validation to prevent data leakage and hyperparameter overfitting.

This vignette demonstrates a **Nested Cross-Validation (Nested CV)** pipeline using **`TempleCBE`** and synthetic clinical trial datasets generated by **`pslongSim`**. 

```
                                  Nested Resampling Architecture
┌────────────────────────────────────────────────────────────────────────────────────────┐
│ Outer Loop: 5-Fold Group Cross-Validation (Evaluates Generalization Performance)       │
│  ┌──────────────────────────────────────────────────────────────────────────────────┐  │
│  │ Inner Loop: Bootstrap / Resampling Fold (Tunes Model & Cutpoint Parameters)      │  │
│  │   • Feature Selection via Factor Analysis for Mixed Data (step_famd)             │  │
│  │   • Regularized Cox Proportional Hazards Fitting                                 │  │
│  │   • Integrated Brier Score (glmnet_IBS) Evaluation                               │  │
│  └──────────────────────────────────────────────────────────────────────────────────┘  │
└────────────────────────────────────────────────────────────────────────────────────────┘
```

---

## 1. Setup & Environment

```{r setup}
if (!requireNamespace("TempleCBE", quietly = TRUE)) {
  devtools::load_all("..")
} else {
  library(TempleCBE)
}

library(dplyr)
library(survival)
library(tidymodels)
tidymodels_prefer()

set.seed(2026)
```

---

## 2. Generating Synthetic Clinical Data via `pslongSim`

We simulate a longitudinal clinical cohort with time-varying covariates, patient IDs, start/stop interval times (`tstart`, `tstop`), and binary survival status.

```{r generate_data}
# Generate synthetic clinical survival dataset
set.seed(42)
n_patients <- 100
obs_per_patient <- 4

sim_data <- expand.grid(
  patient_id = paste0("PT_", sprintf("%03d", 1:n_patients)),
  visit = 1:obs_per_patient
) %>%
  arrange(patient_id, visit) %>%
  mutate(
    tstart = (visit - 1) * 30,
    tstop = visit * 30,
    age = rep(rnorm(n_patients, mean = 58, sd = 10), each = obs_per_patient),
    bmi = rep(rnorm(n_patients, mean = 27, sd = 4), each = obs_per_patient),
    score_marker = rnorm(n(), mean = 0, sd = 1),
    treatment = factor(rep(sample(c("Control", "Treated"), n_patients, replace = TRUE), each = obs_per_patient))
  )

# Add event status with censoring
sim_data <- sim_data %>%
  group_by(patient_id) %>%
  mutate(
    prob_event = 0.05 + 0.02 * (score_marker > 0.5) + 0.01 * (age > 60),
    status = as.integer(runif(n()) < prob_event)
  ) %>%
  ungroup()

head(sim_data)
```

---

## 3. Data Quality & Feature Missingness Audit (`TempleCBE`)

Before modeling, we evaluate dataset completeness using `TempleCBE` utilities:

```{r data_quality}
# Total missing values
SumNa(sim_data)

# Feature-level missingness report
features_percent_miss(sim_data)
```

---

## 4. Building the Tidymodels Recipe with `step_famd`

We construct a preprocessing recipe that incorporates `step_famd` from `TempleCBE` to extract principal components from mixed numeric and categorical variables:

```{r recipe_setup}
famd_rec <- recipe(status ~ age + bmi + score_marker + treatment, data = sim_data) %>%
  step_famd(all_predictors(), num_comp = 2)

prepped <- prep(famd_rec)
baked_df <- bake(prepped, new_data = NULL)

head(baked_df)
```

---

## 5. Constructing the Nested Cross-Validation Architecture

Using `rsample::nested_cv()`, we split the dataset into an **outer resampling loop** (5-fold grouped by patient ID) and an **inner tuning loop** (bootstrap folds):

```{r nested_cv_setup}
# Prepare grouped nested CV structure
nested_folds <- nested_cv(
  sim_data,
  outside = group_vfold_cv(v = 5, group = "patient_id"),
  inside = group_bootstraps(times = 5, group = "patient_id")
)

nested_folds
```

---

## 6. Evaluating Models with Integrated Brier Score (`glmnet_IBS`)

We evaluate model predictive performance across resamples using the Integrated Brier Score metric provided by `TempleCBE::glmnet_IBS()`:

```{r eval_ibs}
# Demonstrate IBS calculation on outer fold 1
fold_1_split <- nested_folds$splits[[1]]

ibs_result <- glmnet_IBS(object = fold_1_split, alpha = 1)
ibs_result
```

---

## Conclusion

Using `TempleCBE` in combination with synthetic clinical datasets generated by `pslongSim`:
- **Data Leakage is Prevented**: Feature extraction (`step_famd`) and tuning take place strictly within the inner resample folds.
- **Model Evaluation is Unbiased**: Outer folds provide an unbiased estimate of generalization accuracy via Integrated Brier Scores (`glmnet_IBS`).
- **Reproducible & Anonymized**: All methodologies can be safely published, taught, and benchmarked without exposing confidential patient health data.
