---
title: "Penalized Cox Models and Nested Cross-Validation for Start/Stop Survival Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Penalized Cox Models and Nested Cross-Validation for Start/Stop Survival Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  dev = "png"
)
required <- c("glmnet", "survival", "rsample", "yardstick", "recipes", "tidymodels")
knitr::opts_chunk$set(eval = all(vapply(required, requireNamespace, logical(1), quietly = TRUE)))
```

## Why start/stop data needs care

In counting-process (start/stop) layout, each subject has one row per
interval, `Surv(tstart, tstop, status)`, so covariates can change over
follow-up. Three things go wrong if those rows are treated as independent
observations:

1. **Resampling leaks.** Row-level folds, like those of `glmnet::cv.glmnet()`,
   put some of a subject's intervals in the analysis set and others in the
   assessment set.
2. **Scoring is wrong, silently.** yardstick's survival metrics expect one
   row per subject with right-censored truth. Given start/stop truth, they
   return numbers without complaint, but those numbers count every interval as
   its own subject.
3. **Censoring is ignored.** Brier scores need inverse-probability-of-censoring
   weights, or subjects lost to follow-up bias the score.

`TempleCBE` handles each of these:

| Function | What it does |
|---|---|
| `coxnet()` | Elastic-net Cox model (glmnet) with formula, recipe, and x/y interfaces (hardhat) |
| `cv_coxnet()` | Chooses `mixture` and `penalty` on subject-grouped folds with a yardstick metric set |
| `nested_cv_coxnet()` | Estimates how well that whole tuning procedure generalizes |
| `surv_subject_truth()`, `censoring_km()`, `graf_weights()`, `add_graf_weights()` | Score any start/stop model with yardstick |

```{r setup}
library(TempleCBE)
library(tidymodels)
library(survival)
tidymodels_prefer()
```

## Simulated data

Each patient has one to four visits at irregular times, belongs to one of five
sites, and has a marker that drifts between visits. An event can only occur
in a patient's last interval.

```{r generate_data}
set.seed(42)
sim_data <- purrr::map(seq_len(150), function(i) {
  n_visits <- sample(1:4, 1)
  tstop <- round(cumsum(runif(n_visits, 20, 40)))
  age <- rnorm(1, mean = 58, sd = 10)
  treatment <- sample(c("Control", "Treated"), 1)
  marker <- rnorm(1) + cumsum(rnorm(n_visits, sd = 0.3))
  risk <- 0.03 * (age - 58) + 0.8 * marker[n_visits] - 0.6 * (treatment == "Treated")
  tibble(
    patient_id = sprintf("PT_%03d", i),
    site = paste0("site_", i %% 5 + 1),
    tstart = c(0, head(tstop, -1)),
    tstop = tstop,
    status = c(rep(0L, n_visits - 1), rbinom(1, 1, plogis(risk))),
    age = age,
    bmi = rnorm(1, mean = 27, sd = 4),
    marker = marker,
    treatment = factor(treatment, levels = c("Control", "Treated"))
  )
}) |>
  purrr::list_rbind()

sim_data
```

Collapsed to one row per patient, this is the truth yardstick needs:

```{r truth}
surv_subject_truth(Surv(sim_data$tstart, sim_data$tstop, sim_data$status), sim_data$patient_id)
```

## Fitting one model: `coxnet()`

`coxnet()` fits the whole glmnet regularization path. The formula interface
expands factors into indicator columns.

```{r coxnet}
fit <- coxnet(
  Surv(tstart, tstop, status) ~ age + bmi + marker + treatment,
  data = sim_data, mixture = 0.5, penalty = 0.01
)
fit
tidy(fit)
```

Predictions follow tidymodels conventions: `.pred_linear_pred` (larger means
longer survival), or a `.pred` list-column of survival probabilities.

```{r coxnet_predict}
predict(fit, sim_data[1:3, ], type = "survival", eval_time = c(30, 60, 90))$.pred[[1]]
```

## Choosing `mixture` and `penalty`: `cv_coxnet()`

With a recipe, preprocessing is learned inside each fold. The outcome is a
`Surv` column; the patient identifier gets the `"id"` role, which
`cv_coxnet()` uses to group folds and score patients.

```{r cv_coxnet}
sim_data$surv <- Surv(sim_data$tstart, sim_data$tstop, sim_data$status)

cox_rec <- recipe(surv ~ age + bmi + marker + treatment + patient_id + site, data = sim_data) |>
  update_role(patient_id, new_role = "id") |>
  update_role(site, new_role = "site") |>
  step_dummy(treatment) |>
  step_normalize(all_numeric_predictors())

set.seed(2026)
cv <- cv_coxnet(cox_rec, sim_data, mixture = c(0.25, 0.5, 1), v = 5, nlambda = 30)
cv
```

By default every fold is scored with the integrated Brier score (used to
choose), concordance, and the time-specific Brier score and ROC AUC, at deciles
of the observed event times.

```{r cv_metrics}
collect_metrics(cv) |>
  filter(.metric %in% c("brier_survival_integrated", "concordance_survival")) |>
  arrange(.metric, mean)
autoplot(cv)
tidy(cv, penalty = "lambda.1se")
```

Folds can be grouped by a coarser unit than the patient. With
`group = "site"`, each fold holds out whole sites:

```{r cv_sites}
set.seed(2026)
cv_sites <- cv_coxnet(
  cox_rec, sim_data, group = "site", v = 5, nlambda = 20,
  metrics = metric_set(brier_survival_integrated, concordance_survival)
)
cv_sites$best_by_mixture
```

## How well does tuning generalize? `nested_cv_coxnet()`

A cross-validated score at the chosen settings is optimistic: the same
assessment sets chose the settings. Nested cross-validation tunes on inner
resamples, refits on each outer analysis set, and scores the untouched outer
assessment set.

```{r nested}
set.seed(2026)
nested_folds <- nested_cv(
  sim_data,
  outside = group_vfold_cv(group = patient_id, v = 3),
  inside = group_vfold_cv(group = patient_id, v = 3)
)

nested <- nested_cv_coxnet(
  nested_folds, cox_rec,
  mixture = c(0.5, 1), rule = "1se", nlambda = 20,
  metrics = metric_set(brier_survival_integrated, concordance_survival)
)
nested |> select(id, mixture, penalty)
collect_metrics(nested)
```

## Scoring any start/stop model with yardstick

The same helpers score predictions from any model. Here, a `coxnet()` fit
predicts each test patient's survival from their baseline covariates:

```{r any_model}
set.seed(2026)
split <- group_initial_split(sim_data, group = patient_id)
train <- training(split)
test <- testing(split)
times <- c(30, 60, 90)

model <- coxnet(
  Surv(tstart, tstop, status) ~ age + bmi + marker + treatment,
  data = train, mixture = cv$mixture, penalty = cv$lambda_min
)

train_truth <- surv_subject_truth(train$surv, train$patient_id)
test_truth <- surv_subject_truth(test$surv, test$patient_id)
baseline_rows <- test |> filter(tstart == 0) |> arrange(patient_id)

scored <- test_truth |>
  bind_cols(predict(model, baseline_rows, type = "survival", eval_time = times)) |>
  add_graf_weights(censoring = censoring_km(train_truth$.truth))

brier_survival_integrated(scored, truth = .truth, .pred)
```

## Existing analysis code: `glmnet_IBS()`

`glmnet_IBS()`, `tune_over_alpha()`, and `summarize_tune_results()` keep their
arguments. They now use `cv_coxnet()` underneath, so the penalty is chosen on
patient-grouped folds and the IBS uses censoring weights.

```{r glmnet_ibs}
legacy_rec <- recipe(~ age + bmi + marker + treatment + patient_id + tstart + tstop + status, data = sim_data) |>
  update_role(patient_id, tstart, tstop, status, new_role = "id variable") |>
  step_dummy(treatment) |>
  step_normalize(all_numeric_predictors())

set.seed(2026)
glmnet_IBS(
  split, alpha = 0.5, recipe = legacy_rec,
  feature_names = c("age", "bmi", "marker", "treatment_Treated"),
  eval_time = times, id_col = "patient_id", internal_folds = 3
)
```
