---
title: "06. Python Interoperability: Validating LOCO-MP Survival Ensembles with reticulate, rpy2, and DataSlingers/LOCOMP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{06. Python Interoperability: Validating LOCO-MP Survival Ensembles with reticulate, rpy2, and DataSlingers/LOCOMP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup_knitr, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  dev = "png",
  dpi = 150,
  fig.width = 7,
  fig.height = 4.5
)
```

```{r load_libraries}
library(TempleCBE)
library(survival)
library(dplyr)
library(tibble)
library(ggplot2)

# Check Python and reticulate availability
has_reticulate <- requireNamespace("reticulate", quietly = TRUE)
has_python <- FALSE
if (has_reticulate) {
  has_python <- tryCatch(
    reticulate::py_available(initialize = TRUE),
    error = function(e) FALSE
  )
}
```

## Introduction: Model-Agnostic Feature Importance Inference

Quantifying the predictive contribution of individual covariates is central to biomedical discovery, translational biomarker prioritization, and regulatory evaluation. Traditional machine learning interpretability methods face substantial limitations:

1. **Permutation Feature Importance (PFI)**: Permuting a single feature can create unphysical combinations in the presence of strong collinearity, inflating false-positive discovery.
2. **Drop-Column (Retraining) LOCO**: Training $p$ distinct models, each omitting one covariate, is computationally prohibitive on high-dimensional genomic arrays and complex survival ensembles.
3. **Absence of Formal Uncertainty Quantification**: Popular attribution tools (such as tree-based feature importances or standard SHAP values) yield point estimates without distribution-free hypothesis tests, standard errors, or valid confidence intervals.

To resolve these challenges, **Gan, Zheng, & Allen (2022)** introduced **LOCO-MP** (*Leave-One-Covariate-Out Inference via MiniPatch Ensembles*, [arXiv:2206.02088](https://arxiv.org/abs/2206.02088)), implemented in Python at [`DataSlingers/LOCOMP`](https://github.com/DataSlingers/LOCOMP). 

This vignette demonstrates:
- How `TempleCBE::cbe_loco_mp_coxnet()` translates the Python `LOCOMP` framework into **censored survival analysis** using Inverse Probability of Censoring Weighting (IPCW) Graf Integrated Brier Score loss.
- How to orchestrate bidirectional cross-language pipelines using **`reticulate`** (calling Python from R) and **`rpy2`** (calling `TempleCBE` from Python).
- Empirical validation and feature ranking concordance on simulated clinical cohorts.

---

## Part 1: The LOCO-MP Algorithmic Framework

LOCO-MP builds an ensemble of $B$ random **minipatches**, where each patch simultaneously subsamples both observations and features without replacement:

$$\text{Patch } b = (\mathcal{S}_b, \mathcal{F}_b), \quad |\mathcal{S}_b| = \lfloor n \cdot n_{\text{ratio}} \rfloor, \quad |\mathcal{F}_b| = \lfloor p \cdot m_{\text{ratio}} \rfloor$$

Typical defaults are $n_{\text{ratio}} = 0.7$ (70% of subjects) and $m_{\text{ratio}} = 0.5$ (50% of candidate features).

```
   Original Data (N x P)              B Minipatches (N_sub x P_sub)
┌───────────────────────────┐         ┌───────────┐   ┌───────────┐
│                           │  ====>  │  Patch 1  │...│  Patch B  │
│                           │         └───────────┘   └───────────┘
└───────────────────────────┘               │               │
                                            ▼               ▼
                                      Fit Model 1     Fit Model B
                                            │               │
                                            ▼               ▼
                                 Out-of-Bag (OOB) Prediction
                                     Full vs Leave-One-Out
                                            │
                                            ▼
                             Distribution-Free Inference:
                             SE, Z-score, CI, Adjusted p
```

### Out-of-Bag Leave-One-Out Evaluation

For each subject $i$, predictions are formed strictly **out-of-bag** (OOB) using only patches where subject $i$ was *not* sampled ($i \notin \mathcal{S}_b$):

- **$\hat{y}_{i, +j}$ (Full Ensemble)**: Aggregate prediction across OOB patches containing feature $j$ ($j \in \mathcal{F}_b$).
- **$\hat{y}_{i, -j}$ (Leave-One-Out Ensemble)**: Aggregate prediction across OOB patches omitting feature $j$ ($j \notin \mathcal{F}_b$).

The individual prediction loss difference is defined as:

$$D_{i,j} = L(y_i, \hat{y}_{i, -j}) - L(y_i, \hat{y}_{i, +j})$$

- If $D_{i,j} > 0$, omitting feature $j$ increases prediction error; feature $j$ carries genuine predictive signal.
- If $D_{i,j} \approx 0$ or $D_{i,j} < 0$, feature $j$ is non-informative or redundant.

---

## Part 2: Mathematical Parity: Python `DataSlingers/LOCOMP` vs. R `TempleCBE`

While the core minipatch architecture is identical, survival analysis introduces censoring and time-dependence that require distinct scoring metrics:

| Attribute | Python `DataSlingers/LOCOMP` (`LOCOMPReg`) | R `TempleCBE` (`cbe_loco_mp_coxnet`) |
|:---|:---|:---|
| **Target Outcome ($Y$)** | Continuous scalar ($y \in \mathbb{R}$) | Right-censored `Surv(time, status)` or start/stop intervals |
| **Loss Function ($L$)** | Mean Squared Error: $(y - \hat{y})^2$ | IPCW Graf Integrated Brier Score loss |
| **Base Estimator** | Ridge, SVR, Random Forest | Penalized Cox Proportional Hazards (`coxnet`) |
| **Longitudinal Data** | Independent row records | Grouped subject clustering (`id = id`) |
| **Censoring Adjustment** | None (uncensored) | Kaplan-Meier IPCW weights $\hat{G}(t) = P(C > t)$ |
| **Inference Statistics** | Asymptotic normal $Z$-test, Bonferroni | Asymptotic normal $Z$-test, Bonferroni, BH, FDR |

### The IPCW Survival Loss Formulation in TempleCBE

In survival models, squared error cannot be calculated directly because true event times are censored for patients who remain event-free at follow-up. `TempleCBE` computes the **IPCW Graf Integrated Brier Loss** over an evaluation grid $\tau_1 < \tau_2 < \dots < \tau_K$:

$$L(i, \hat{S}) = \frac{1}{\tau_K - \tau_1} \sum_{k=1}^{K-1} (\tau_{k+1} - \tau_k) \cdot \ell\left(T_i, \delta_i, \hat{S}(\tau_k \mid x_i), \tau_k\right)$$

where the IPCW loss at time $\tau_k$ is:

$$\ell(T_i, \delta_i, \hat{S}_i, \tau_k) = \left(0 - \hat{S}_i(\tau_k)\right)^2 \frac{I(T_i \le \tau_k, \delta_i = 1)}{\hat{G}(T_i)} + \left(1 - \hat{S}_i(\tau_k)\right)^2 \frac{I(T_i > \tau_k)}{\hat{G}(\tau_k)}$$

### Asymptotic Statistical Inference

Both the Python and R implementations compute the sample mean importance and empirical variance:

$$\bar{D}_j = \frac{1}{n} \sum_{i=1}^n D_{i,j}, \qquad \widehat{\text{SE}}(\bar{D}_j) = \sqrt{\frac{1}{n(n-1)} \sum_{i=1}^n (D_{i,j} - \bar{D}_j)^2}$$

As proved in Theorem 1 of Gan et al. (2022), under mild regularity conditions, the test statistic converges asymptotically to a standard normal distribution:

$$Z_j = \frac{\bar{D}_j}{\widehat{\text{SE}}(\bar{D}_j)} \xrightarrow{d} \mathcal{N}(0, 1) \quad \text{under } H_0: \mathbb{E}[D_{i,j}] \le 0$$

The one-sided $p$-value and $(1 - \alpha)$ simultaneous confidence intervals are:

$$p_j = 1 - \Phi(Z_j), \qquad \text{CI}_{1-\alpha}(j) = \left[ \bar{D}_j - z_{1-\alpha/(2p)} \widehat{\text{SE}}(\bar{D}_j), \; \bar{D}_j + z_{1-\alpha/(2p)} \widehat{\text{SE}}(\bar{D}_j) \right]$$

---

## Part 3: Workflow A: Calling Python `DataSlingers/LOCOMP` from R via `reticulate`

In hybrid scientific teams, statistical analysts often evaluate continuous biomarker targets in Python using `DataSlingers/LOCOMP`. `reticulate` provides bidirectional data sharing between R and Python without intermediate file serialization.

### Python Environment Verification

```{r verify_python}
if (has_python) {
  py_info <- reticulate::py_config()
  cat("Active Python:", as.character(py_info$version), "\n")
  cat("Python Path:  ", py_info$python, "\n")
} else {
  cat("Python is not currently available in this environment. Showing reproducible workflow syntax.\n")
}
```

### Calling Python LOCO-MP from R

The code below demonstrates running `DataSlingers/LOCOMP` from R:

```r
# In R: Import Python modules via reticulate
library(reticulate)

# Import DataSlingers LOCOMP package
locomp <- import("locomp")
ml_models <- import("ML_models")

# Convert R data frame to NumPy arrays
X_py <- np_array(as.matrix(my_data[, feature_names]))
y_py <- np_array(my_data$continuous_outcome)

# Fit LOCO-MP regression ensemble in Python
loco_fit <- locomp$LOCOMPReg(
  X = X_py,
  Y = y_py,
  n_ratio = 0.7,
  m_ratio = 0.5,
  B = 100L,
  fit_func = ml_models$ridge2,
  alpha = 0.05,
  bonf = TRUE
)

# Access Python results directly in R
py_results <- data.frame(
  term       = feature_names,
  importance = as.numeric(loco_fit$info$mean_diff),
  std_error  = as.numeric(loco_fit$info$se),
  z_stat     = as.numeric(loco_fit$info$z_score),
  p_value    = as.numeric(loco_fit$info$p_val)
)
```

---

## Part 4: Workflow B: Calling `TempleCBE` from Python via `rpy2`

Python clinical data scientists working with survival frameworks (such as `scikit-survival` or `lifelines`) can call `TempleCBE` directly through `rpy2` to perform LOCO-MP survival inference.

### Python Script Calling TempleCBE

```python
"""
Python script: Call TempleCBE LOCO-MP Cox model via rpy2
"""
import numpy as np
import pandas as pd
import rpy2.robjects as ro
from rpy2.robjects import pandas2ri
from rpy2.robjects.packages import importr

# Activate automated Pandas <-> R DataFrame conversion
pandas2ri.activate()

# Import TempleCBE and survival packages
templecbe = importr('TempleCBE')
survival = importr('survival')
base = importr('base')

# 1. Load or simulate survival data in Python
n = 120
data_py = pd.DataFrame({
    'time': np.random.exponential(scale=10, size=n) + 1,
    'status': np.random.binomial(1, 0.65, size=n),
    'age': np.random.normal(55, 10, size=n),
    'biomarker_a': np.random.normal(0, 1, size=n),
    'biomarker_b': np.random.normal(0, 1, size=n),
    'noise_1': np.random.normal(0, 1, size=n),
    'noise_2': np.random.normal(0, 1, size=n)
})

# 2. Convert Pandas DataFrame to R DataFrame
data_r = pandas2ri.py2rpy(data_py)

# 3. Fit LOCO-MP Penalized Cox Model in TempleCBE
fit_r = templecbe.cbe_loco_mp_coxnet(
    formula = ro.Formula("Surv(time, status) ~ age + biomarker_a + biomarker_b + noise_1 + noise_2"),
    data = data_r,
    B = 50,
    n_ratio = 0.7,
    m_ratio = 0.5,
    seed = 42
)

# 4. Extract tidy results back into Pandas
generics = importr('generics')
tidy_df_r = generics.tidy(fit_r)
tidy_df_py = pandas2ri.rpy2py(tidy_df_r)

print("--- TempleCBE LOCO-MP Results in Python ---")
print(tidy_df_py[['term', 'importance', 'std_error', 'statistic', 'p_adjusted', 'conf_low', 'conf_high']])
```

This delivers full biostatistical capabilities to Python pipelines without re-implementing complex Kaplan-Meier IPCW integration routines.

---

## Part 5: Direct Empirical Simulation & Concordance Benchmark

To demonstrate `cbe_loco_mp_coxnet()` in action, we simulate a synthetic cohort where true risk is driven by features $X_1$ and $X_2$, while $X_3, \dots, X_6$ are independent uninformative noise:

$$\log h(t \mid X) = \log h_0(t) + 1.2 X_1 - 0.8 X_2 + 0 X_3 + 0 X_4 + 0 X_5 + 0 X_6$$

### Simulation Setup

```{r simulate_data}
set.seed(42)
n <- 120
p <- 6

# Design matrix with 6 standardized predictors
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("X", 1:p)

# True linear predictor: X1 and X2 carry real predictive signal
beta_true <- c(1.2, -0.8, 0, 0, 0, 0)
risk <- as.vector(X %*% beta_true)

# Simulate survival times under proportional hazards
time <- rexp(n, rate = exp(risk))
status <- rbinom(n, 1, prob = 0.70)

sim_cohort <- as.data.frame(X)
sim_cohort$time <- time
sim_cohort$status <- status

cat("Simulated clinical cohort: N =", nrow(sim_cohort), "patients, P =", p, "covariates.\n")
cat("Overall event rate: ", round(mean(status) * 100, 1), "%\n")
```

### Running `cbe_loco_mp_coxnet()`

```{r fit_loco_mp}
# Fit LOCO-MP ensemble across B = 50 minipatches
fit_loco <- cbe_loco_mp_coxnet(
  formula   = Surv(time, status) ~ X1 + X2 + X3 + X4 + X5 + X6,
  data      = sim_cohort,
  B         = 50,
  n_ratio   = 0.7,
  m_ratio   = 0.5,
  alpha     = 0.05,
  p_adjust  = "bonferroni",
  seed      = 123
)

# Print executive summary
print(fit_loco)
```

### Tidy Feature Importance Table

We extract the tidy results tibble via `generics::tidy()`:

```{r tidy_loco_results}
loco_table <- generics::tidy(fit_loco)
loco_table %>%
  select(term, importance, std_error, statistic, p_value, p_adjusted, conf_low, conf_high)
```

Notice the clear statistical separation:
- **True Signal Features ($X_1, X_2$)**: Display positive importance values, large positive $Z$-statistics, and Bonferroni-adjusted $p$-values that reach statistical significance.
- **Noise Features ($X_3 \dots X_6$)**: Display negative or near-zero importance values, non-significant $p$-values ($p_{\text{adj}} \approx 1.0$), and confidence intervals spanning zero.

### Visualization with `autoplot()`

`TempleCBE` provides an automated `autoplot()` method generating a publication-ready forest / lollipop plot styled with Temple CBE brand guidelines:

```{r plot_loco_importance}
autoplot(fit_loco, title = "LOCO-MP Feature Importance: Simulated Survival Cohort")
```

---

## Part 6: Cross-Language Data Integrity Verification with `cbe_compare_df()`

When exchanging feature attribution tables between R and Python environments, biostatisticians must verify numerical concordance. `TempleCBE::cbe_compare_df()` serves as an audit-ready reconciliation engine:

```{r reconcile_features}
# Reconcile feature importance table against reference
loco_reference <- loco_table

# Introduce slight perturbation to test reconciliation sensitivity
loco_perturbed <- loco_table
loco_perturbed$importance[1] <- loco_perturbed$importance[1] + 1e-5

cmp_features <- cbe_compare_df(
  base         = loco_reference,
  compare      = loco_perturbed,
  by           = "term",
  tolerance    = 1e-4,
  base_name    = "R_Engine",
  compare_name = "Python_Reconciled"
)

print(cmp_features)
cat("Concordance within tolerance (1e-4):", cmp_features$is_concordant, "\n")
```

---

## Conclusion

By combining the distribution-free inference principles of **`DataSlingers/LOCOMP`** with **IPCW survival theory**, `TempleCBE` provides:

1. **Model-Agnostic Feature Importance for Survival Models**: Valid standard errors, $Z$-tests, and simultaneous confidence intervals without full-model refitting.
2. **Seamless Bidirectional Interoperability**: Direct bridges for both R calling Python (`reticulate`) and Python calling R (`rpy2`).
3. **Audit-Ready Validation**: Automated cell-by-cell discrepancy reporting via `cbe_compare_df()`.
