## Pairwise $\chi^2$ Tests

Every pairwise combination of the categorical variables below is tested; this is exploratory and the resulting p-values are not adjusted for multiplicity, so they should be interpreted accordingly rather than as independent confirmatory findings. A subset of these variables are themselves derived from the same underlying angle-of-impact measurements -- "Perpendicular to" and the "Chest Wall Angle Category"/"Spine Angle Category" variables are all constructed from the same trajectory-angle data (see the Summary Statistics section above for how the angle categories are defined) -- so an association found between those specific variables reflects that shared construction rather than an independent finding. Additionally, the "Unknown" level present in some of these categorical variables represents missing/unresolved data rather than a substantive category; it is included in the tests below as-is.

```{r}
ABG_Biomechanics_cat <- ABG_Biomechanics %>%
    select(-arl_number, -pp_chest_wall, -pp_to_spine, -pp_neither) %>%
    select(-where(is.numeric)) %>%
    select(where(is.character) | where(is.factor)) %>%
    colnames()

cat_combos <- combn(ABG_Biomechanics_cat, 2)

cat_combos <- as.data.frame(cat_combos) %>%
    pivot_longer(everything(), names_to = "match", values_to = "name") %>%
    arrange(match) %>%
    mutate(id = stringr::str_remove_all(match, "V")) %>%
    mutate(id = as.numeric(id)) |>
    left_join(new_labels_df, join_by(name == variable))
```

```{r}
cat_ids <- cat_combos %>%
    select(id) %>%
    distinct() %>%
    pull()
```

```{r}
TT <- function(x) {
    cur_combo <- cat_combos %>%
        filter(id == x) %>%
        select(name) %>%
        pull()

    cur_combo

    ABG_Biomechanics %>%
        select(all_of(cur_combo)) %>%
        table()
}
```

```{r}
GG <- function(x) {
    X <- chisq.test(TT(x), correct = FALSE)

    cur_combo <- names(dimnames(TT(x)))

    cur_combo <- new_labels_df |>
        filter(variable %in% cur_combo) |>
        pull(label)

    p <- ggpubr::ggballoonplot(
        as.data.frame(X$observed),
        main = paste(cur_combo, collapse = " X "),
        size.range = c(5, 20),
        fill = "Freq",
        show.label = TRUE,
        font.label = list(color = "white"),
        caption = paste0(
            "Chi-Square Test: ",
            scales::label_pvalue(
                prefix = c("p < ", "p = ", "p > "),
                add_p = TRUE
            )(X$p.value)
        )
    ) +
        ggpubr::gradient_fill(c("blue", "red")) +
        guides(size = FALSE)
    return(p)
}
```

```{r}
#| results: asis
if (!interactive()) {
    src <- lapply(cat_ids, function(x) {
        knitr::knit_expand('child_eda_chi_square.qmd')
    })
    res <- knitr::knit_child(text = unlist(src), quiet = TRUE)
    cat(res, sep = '\n')
}
```
