---
title: "Example"
geometry: margin=1in
author:
- name: J Kyle Armstrong, PhD
  orcid: 0000-0002-4269-0121
  email: j.kyle.armstrong@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
- name: Author 2
  email: example@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
date: today
date-format: medium
fontsize: 10pt
bibliography:
  - bib.bib
  - grateful-refs.bib
format:
  html:
    smooth-scroll: true
    theme:
      light: flatly
      dark: darkly
    toc: true
    lot: true
    lof: true
    number-depth: 6
    number-sections: true
    section-divs: true
    toc-location: left
    toc-expand: true
  docx:
    toc: true
    lot: true
    lof: true
    number-depth: 6
    number-sections: true
  pdf:
    toc: true
    lot: true
    lof: true
    number-depth: 8
    toc-depth: 4
    standalone: true
    number-sections: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
---


```{r setup, echo= FALSE, message=FALSE}
knitr::opts_chunk$set(include = TRUE) 
knitr::opts_chunk$set(echo = FALSE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 
```

```{r}
library('ggplot2')
library('purrr')
library('dplyr')
library('gtsummary')
library('gt')
```

```{r}
# Self-contained synthetic data: this example ships with the public package
# and must not depend on any private data source.
set.seed(42)
n <- 60
input_parameters_raw <- tibble::tibble(
  arl_number = sprintf("SYN-%03d", seq_len(n)),
  head_shape = sample(c("Round", "Oval"), n, replace = TRUE),
  orientation = sample(c("Frontal", "Lateral", "Occipital"), n, replace = TRUE),
  nom_velocity_m_sec = round(rnorm(n, mean = 5.5, sd = 0.8), 2),
  init_strike_velocity_m_s = round(rnorm(n, mean = 5.2, sd = 0.9), 2),
  depth_mm = round(rgamma(n, shape = 4, rate = 1), 2),
  max_depth_mm = round(depth_mm + rgamma(n, shape = 2, rate = 1), 2),
  duration_ms = round(rgamma(n, shape = 6, rate = 0.8), 2),
  transf_energy_j = round(rgamma(n, shape = 5, rate = 1.2), 2),
  pp_chest_wall = sample(c(0, 1), n, replace = TRUE, prob = c(0.7, 0.3)),
  pp_to_spine = sample(c(0, 1), n, replace = TRUE, prob = c(0.85, 0.15)),
  pp_neither = sample(c(0, 1), n, replace = TRUE, prob = c(0.6, 0.4))
)
```

# Table One

```{r}
#| label: tbl-table1
#| tbl-cap: An example of a Table 1

input_parameters_raw %>%
  select(-arl_number, -pp_chest_wall, -pp_to_spine, -pp_neither) %>%
  tbl_summary(by = "head_shape",
              type = all_continuous() ~ "continuous2",
              statistic = list(
                all_continuous() ~ c("{mean} ({sd})","{median} ({p25}, {p75})","{median} ({min} - {max})")
                )
  ) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2)) %>%
  add_overall(last = TRUE, col_label = "**Overall**") %>%
  separate_p_footnotes() %>% 
  as_gt() %>%
  gt::as_latex() # for pdf documents 
```


## Cross Refrence 
You can cross-reference tables, listings, figures and page numbers 

You can use the labels you created to reference the tables like @tbl-table1 or items in the `bib.bib` file such as @Al-Helali2021-as. 

You can even reference later figures such as @fig-2 or sections like @sec-software or even the page number that something happens on \pageref{fig-3}.


## Pairwise t-tests

Suppose we wanted to visualize all pairwise t-tests (we will only do first four but you get the idea).


```{r}
features_numeric <- input_parameters_raw %>%
  select(-arl_number) %>%
  select(where(is.numeric)) %>%
  colnames()


pairs <- combn(features_numeric,2) |> 
  t() %>%
  as_tibble() %>%
  mutate(pair_number = row_number()) 

pair_number_list <- pairs %>% filter(pair_number <= 4) %>% pull(pair_number)
```




```{r}
#| label: t_test_child
#| results: 'asis'

if(!interactive()){
  src <- lapply(pair_number_list,
                function(pair_number){knitr::knit_expand('t_test_child.qmd')})
  res = knitr::knit_child(text = unlist(src), quiet = TRUE)
  cat(res, sep = '\n')
}
```


\pagebreak
# Software {#sec-software}

```{r}
grateful::cite_packages(output = "paragraph",
                            out.dir = here::here('inst','templates'),
                            omit = NULL,
                            include.RStudio = F,
                            passive.voice = T)
```


# Refrences 