---
title: "Example"
geometry: margin=1in
author:
- name: J Kyle Armstrong, PhD
  orcid: 0000-0002-4269-0121
  email: johnathon.armstrong@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
- name: Example Author 
  email: example@temple.edu
  affiliation: Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology
date: today
date-format: medium
fontsize: 10pt
bibliography:
  - bib.bib
  - grateful-refs.bib
format:
  html:
    smooth-scroll: true
    theme:
      light: flatly
      dark: darkly
    toc: true
    lot: true
    lof: true
    number-depth: 6
    number-sections: true
    section-divs: true
    toc-location: left
    toc-expand: true
  docx:
    toc: true
    lot: true
    lof: true
    number-depth: 6
    number-sections: true
  pdf:
    toc: true
    lot: true
    lof: true
    number-depth: 8
    toc-depth: 4
    standalone: true
    number-sections: true
    template-partials:
      - title.tex
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
        \usepackage{lscape}
        \newcommand{\blandscape}{\begin{landscape}}
        \newcommand{\elandscape}{\end{landscape}}
---


```{r setup, echo= FALSE, message=FALSE}
knitr::opts_chunk$set(include = TRUE) 
knitr::opts_chunk$set(echo = FALSE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 
```

```{r}
library('ggplot2')
library('datasci')
library('purrr')
library('dplyr')
library('gtsummary')
library('gt')
```

```{r}
data(input_parameters_raw)
```

\pagebreak

# Quarto Example 

This is an example of a Quarto document, you will need to install Quarto from <https://quarto.org/>. This package comes with several common R packages, it is also recommended to install:  

```{r}
#| echo: TRUE
#| eval: FALSE

install.packages(c('gt', # general table formatting
                   'gtsummary', # for univariate and regression tables
                   'here', # where you are working (better when working with git)
                   'knitr', # r package for working with documents 
                   'quarto', # R package for working with quarto 
                   'rmarkdown', # creating documents
                   'tidymodels', # modern interface for fitting models 
                   'tidyverse', # modern R language 
                   'tinytex', # needed for pdf documents 
                   'devtools' # package creation development tools
                   ))


remotes::install_github("Pakillo/grateful") # for package citation
```

To create pdf documents it is also recommended to run

```{r}
#| echo: TRUE
#| eval: FALSE

tinytex::install_tinytex()
```

```{r}
#| label: tbl-knitr-example
#| tbl-cap: first 5 records and 5 columns of the data 

input_parameters_raw %>%
  head(n = 5) %>%
  select(arl_number, head_shape, orientation, nom_velocity_m_sec, init_strike_velocity_m_s) %>%
  knitr::kable()
```

# Table One

```{r}
#| label: tbl-table1
#| tbl-cap: "An example of a Table 1"

input_parameters_raw %>%
  select(-arl_number, -pp_chest_wall, -pp_to_spine, -pp_neither) %>%
  tbl_summary(by = "head_shape",
              type = all_continuous() ~ "continuous2",
              statistic = list(
                all_continuous() ~ c("{mean} ({sd})","{median} ({p25}, {p75})","{median} ({min} - {max})")
                )
  ) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2)) %>%
  add_overall(last = TRUE, col_label = "**Overall**") %>%
  separate_p_footnotes() %>% 
  as_gt() %>%
  gt::as_latex() # for pdf documents 
```


# Cross Refrence 
You can cross-reference tables, listings, figures and page numbers 

You can use the `label`s you created to reference the tables like @tbl-table1 or items in the `bib.bib` file such as @Al-Helali2021-as. 

You can even reference later items such as @fig-2 or sections like @sec-software or even the page number that something happens on like page \pageref{fig-3}.


# Pairwise t-tests

Suppose we wanted to visualize all pairwise t-tests (we will only do first four but you get the idea). There are other ways to do this but this is a good enough example of repeated analysis. 


```{r}
features_numeric <- input_parameters_raw %>%
  select(-arl_number) %>%
  select(where(is.numeric)) %>%
  colnames()


pairs <- combn(features_numeric,2) |> 
  t() %>%
  as_tibble() %>%
  mutate(pair_number = row_number()) 

pair_number_list <- pairs %>% filter(pair_number <= 4) %>% pull(pair_number)
```


```{r}
#| label: t_test_child
#| results: 'asis'

if(!interactive()){
  src <- lapply(pair_number_list,
                function(pair_number){knitr::knit_expand('t_test_child.qmd')})
  res = knitr::knit_child(text = unlist(src), quiet = TRUE)
  cat(res, sep = '\n')
}
```


\pagebreak
# Software {#sec-software}

```{r}
grateful::cite_packages(output = "paragraph",
                        out.dir = getwd(),
                        omit = NULL,
                        include.RStudio = F,
                        passive.voice = T)
```


# Refrences 