---
title: "Example"
author:
  - J Kyle Armstrong, PhD^[Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology, j.kyle.armstrong@temple.edu]
  - Example Author^[Lewis Katz School of Medicine at Temple University Center for Biostatistics & Epidemiology, example@temple.edu]
date: "`r format(Sys.time(), '%d-%b-%Y')`"
fontsize: 10pt
bibliography:
  - bib.bib
  - grateful-refs.bib
output:
  bookdown::pdf_document2:
    toc: true
    number_sections: true
  word_document:
    toc: true
    number_sections: true
  html_document:
    toc: true
    number_sections: true
    theme: flatly
---


```{r, echo=FALSE, results='asis'}
#cat('
#\usepackage[noblocks]{authblk}
#\renewcommand*{\Authsep}{, }
#\renewcommand*{\Authand}{, }
#\renewcommand*{\Authands}{, }
#\renewcommand\Affilfont{\small}
#')
```

```{r setup, echo = FALSE, include=TRUE, message=FALSE}
knitr::opts_chunk$set(include = TRUE) 
knitr::opts_chunk$set(echo = FALSE) 
knitr::opts_chunk$set(message = FALSE) 
knitr::opts_chunk$set(warning = FALSE) 
```


```{r}
library('ggplot2')
library('purrr')
library('dplyr')
library('gtsummary')
library('gt')
```

```{r}
# Self-contained synthetic data: this example ships with the public package
# and must not depend on any private data source.
set.seed(42)
n <- 60
example_data <- tibble::tibble(
  subject_id = sprintf("SYN-%03d", seq_len(n)),
  group = sample(c("A", "B"), n, replace = TRUE),
  category = sample(c("X", "Y", "Z"), n, replace = TRUE),
  measure_1 = round(rnorm(n, mean = 5.5, sd = 0.8), 2),
  measure_2 = round(rnorm(n, mean = 5.2, sd = 0.9), 2),
  measure_3 = round(rgamma(n, shape = 4, rate = 1), 2),
  measure_4 = round(measure_3 + rgamma(n, shape = 2, rate = 1), 2),
  measure_5 = round(rgamma(n, shape = 6, rate = 0.8), 2),
  measure_6 = round(rgamma(n, shape = 5, rate = 1.2), 2),
  flag_a = sample(c(0, 1), n, replace = TRUE, prob = c(0.7, 0.3)),
  flag_b = sample(c(0, 1), n, replace = TRUE, prob = c(0.85, 0.15)),
  flag_c = sample(c(0, 1), n, replace = TRUE, prob = c(0.6, 0.4))
)
```

\pagebreak

# Quarto Example 

This is an example of an Rmarkdown. This package comes with several common R packages, it is also recommended to install:  

```{r, echo=TRUE, eval=FALSE}
install.packages(c('gt', # general table formatting
                   'gtsummary', # for univariate and regression tables
                   'here', # where you are working (better when working with git)
                   'knitr', # r package for working with documents 
                   'quarto', # R package for working with quarto 
                   'rmarkdown', # creating documents
                   'tidymodels', # modern interface for fitting models 
                   'tidyverse', # modern R language 
                   'tinytex', # needed for pdf documents 
                   'devtools' # package creation development tools
                   ))


remotes::install_github("Pakillo/grateful") # for package citation
```

To create pdf documents it is also recommended to run

```{r, echo=TRUE, eval=FALSE}
tinytex::install_tinytex()
```

```{r knitr-example, tab.cap="first 5 records and 5 columns of the data"}
example_data %>%
  head(n = 5) %>%
  select(subject_id, group, category, measure_1, measure_2) %>%
  knitr::kable()
```

# Table One

```{r table1, tab.cap="An example of a Table 1"}
example_data %>%
  select(-subject_id, -flag_a, -flag_b, -flag_c) %>%
  tbl_summary(by = "group",
              type = all_continuous() ~ "continuous2",
              statistic = list(
                all_continuous() ~ c("{mean} ({sd})","{median} ({p25}, {p75})","{median} ({min} - {max})")
                )
  ) %>%
  add_p(pvalue_fun = ~ style_pvalue(.x, digits = 2)) %>%
  add_overall(last = TRUE, col_label = "**Overall**") %>%
  separate_p_footnotes() %>% 
  as_gt() %>%
  gt::as_latex() # for pdf documents 
```


# Cross Refrence 
You can cross-reference tables, listings, figures and page numbers 

You can use the `label`s you created to reference the tables like \@ref(tab:table1) or items in the `bib.bib` file such as @Al-Helali2021-as. 

You can even reference later items such as @fig-2 or sections like \@ref(software) or even the page number that something happens on like page \pageref{fig-3}.


# Pairwise t-tests

Suppose we wanted to visualize all pairwise t-tests (we will only do first four but you get the idea). There are other ways to do this but this is a good enough example of repeated analysis. 


```{r}
features_numeric <- example_data %>%
  select(-subject_id) %>%
  select(where(is.numeric)) %>%
  colnames()


pairs <- combn(features_numeric,2) |> 
  t() %>%
  as_tibble() %>%
  mutate(pair_number = row_number()) 

pair_number_list <- pairs %>% filter(pair_number <= 4) %>% pull(pair_number)
```


```{r, label='t_test_child', results='asis'}
if(!interactive()){
  src <- lapply(pair_number_list,
                function(pair_number){knitr::knit_expand('t_test_child.Rmd')})
  res = knitr::knit_child(text = unlist(src), quiet = TRUE)
  cat(res, sep = '\n')
}
```


\pagebreak
# Software {#sec:software}

```{r}
grateful::cite_packages(output = "paragraph",
                        out.dir = getwd(),
                        omit = NULL,
                        include.RStudio = F,
                        passive.voice = T)
```


# Refrences 